---
title: 'Class 4b: Simple Linear Regression'
author: "Business Forecasting"
output:
  xaringan::moon_reader:
    self_contained: true
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: true
      
---   
<style type="text/css">
.remark-slide-content {
    font-size: 20px;
}


</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dpi=300)
library(shiny)
library(ggplot2)
library(forecast)
library(plotly)
library(dplyr)
library(igraph)
library(reshape)
library(spData)
library(leaflet)
library(readr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(hrbrthemes)
library(gridExtra)
library(cowplot)
library(viridis)
library(gapminder)
library(knitr)
library(kableExtra)
library(DT)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", 
colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  blue = "#1E90FF",
  white = "#FFFFFF"
))
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(MASS) # for truehist function
load("Daily_CMX_means2.Rda")

```
---
## Roadmap


### This class
- Testing assumptions behind residuals
  - Linearity
  - Constant Variance
  - Uncorrelated Residuals
  - Normality
  
---
Intro:

what are the main assumptions


Additional assumption
-normality


How can we test them?
formal ways plus eyeballing

---
1. Assumptions behind residuals


- plotting residuals vs fitted values

---

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
set.seed(123)
x <- seq(-5, 5, length.out = 70)
y <- 2 + 2 * x +x^2+ rnorm(70, sd = 3)  # Linear relationship

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
library(ggplot2)
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side

grid.arrange(gg1, gg2, ncol = 2)


```

- Non linearity:
1. graph fittted line to quadratic equation
2. Resulting residuals

 (vs linear one)
 
 
---


```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
set.seed(127)
x <- seq(-5, 5, length.out = 70)
y <- 2 + 2 * x+ rnorm(70, sd = 3)  # Linear relationship

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
library(ggplot2)
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side

grid.arrange(gg1, gg2, ncol = 2)


```

---

- heteroskedasticity (why we need this assumption)


```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data with heteroskedasticity (decreasing variance as x increases)
set.seed(123)
x <- seq(0.1, 5, length.out = 100)
y <- 2 + 3 * x + rnorm(100, sd = 0.9 * (1 - x/5))  # Heteroskedasticity

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side
grid.arrange(gg1, gg2, ncol = 2)

```

---

- heteroskedasticity (why we need this assumption)


```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data with heteroskedasticity (decreasing variance as x increases)
set.seed(123)
x <- seq(0.1, 5, length.out = 100)
y <- 2 + 3 * x + rnorm(100, sd = 0.9)  # Heteroskedasticity

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side
grid.arrange(gg1, gg2, ncol = 2)

```
2 plots, homoskedastic vs heteroskedastic


---

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
# Generate sample data for a linear relationship
set.seed(123)
x <- seq(-5, 5, length.out = 70)
y <- 2 + 2 * x +x^2+ rnorm(70, sd = 3)  # Linear relationship

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_minimal() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_minimal()

# Combine the two plots side by side

grid.arrange(gg1, gg2, ncol = 2)
```
- plitting residuals vs residuals - correlation (dependence)

- one plot with dependence, one without

- not a problem for E(), problem for variance 

---

outlier - interactive one from seeing theory
how to detect it

---
Couple of examples of stuff to test their knowledge


heteroskecasticity - od lewej i prawej higher (albo lower)

linearity - maybe something with logharitm or exp or 1/x
---
```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
set.seed(123)
x <- seq(0.1, 5, length.out = 100)  # Avoid x = 0 for division
y <- 2 + 1 / x + rnorm(100, sd = 0.3)  # Relationship like 1/x

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

gg2
```
---

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
# Function to generate data with specified residual distribution and plot
generate_and_plot_data <- function(residual_dist, df = 100) {
  # Generate data with specified residual distribution
  set.seed(123)
  x <- rnorm(200)
  epsilon <- switch(
    residual_dist,
    "normal" = rnorm(200, sd = 3),
    "exponential" = rexp(200, rate = 1/3),
    "t" = rt(200, df = df)
  )
  y <- 2 + 3 * x + epsilon
  
  # Create a data frame
  data <- data.frame(x, y)
  
  # Fit a linear regression model
  model <- lm(y ~ x, data)
  
  # Create a scatterplot with fitted regression line
  scatterplot <- ggplot(data, aes(x = x, y = y)) +
    geom_point(color = "black") +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(x = "X", y = "Y") +
    theme_minimal()
  
  # Calculate residuals
  residuals <- resid(model)
  
  # Create a histogram of residuals
  histogram <- ggplot(data.frame(Residuals = residuals), aes(x = Residuals)) +
    geom_histogram(binwidth = 0.5, fill = "green", color = "black", alpha = 0.7) +
    labs(x = "Residuals", y = "Frequency") +
    theme_minimal()
  
  # Arrange the scatterplot and histogram side by side
  grid.arrange(scatterplot, histogram, ncol = 2)
}

# Create three sets of scatterplots and histograms with different residual distributions
generate_and_plot_data("normal")

```

---

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
generate_and_plot_data("exponential")

```

---

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
generate_and_plot_data("t", df = 3)
```

---
We will compare the quantiles of residuals to quantiles of normal!



---

- histogram of residuals
```{r, echo=FALSE, results='asis'}
# Create a data frame for the table
# Set the seed for reproducibility
set.seed(123)

# Generate normally distributed residuals with standard deviation = 3
residuals <- round(rnorm(100, sd = 3),3)

# Standardize the residuals
standardized_residuals <- round(scale(residuals),3)

# Create a data frame with the residuals and standardized residuals
data_df <- data.frame(Residuals = residuals, Standardized_Residuals = standardized_residuals)

# Calculate the order (rank) of standardized residuals from smallest to largest
data_df$Index <- rank(data_df$Standardized_Residuals)
data_df$Cumulative_Probability <- round(data_df$Index / (nrow(data_df) + 1),3)
data_df$Quantile_Normal <- round(qnorm(data_df$Cumulative_Probability),3)


# Display the data in a table
datatable(data_df,
          fillContainer = FALSE,
          options = list(
            pageLength = 6,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE
)
```
---
d
- quantile quantile plot

-lighter tail/ heavier tail

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
qq_plot <- ggplot(data_df) +
  geom_qq(aes(sample = Standardized_Residuals)) +
  geom_abline(color = "red") +
  coord_fixed() +
  theme_xaringan()

# Create a histogram with the density of the standard normal distribution overlaid
histogram_plot <- ggplot(data_df, aes(x = Standardized_Residuals)) +
  geom_histogram(binwidth = 0.5, aes(y = after_stat(density)), fill = "green", color = "black", alpha = 0.7) +
  geom_density(color = "blue") +
  labs(x = "Standardized Residuals", y = "Density") +
  theme_xaringan()

# Arrange the QQ plot and histogram side by side
grid.arrange(qq_plot, histogram_plot, ncol = 2)
```



---

- Jarque Bera's normality test for stochastic error

---


WHAT TO DO?

---

Couple of examples to test their knowledge



