---
title: 'Class 4b: Simple Linear Regression diagnostics'
author: "Business Forecasting"
output:
  xaringan::moon_reader:
    self_contained: true
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: true
      
---   
<style type="text/css">
.remark-slide-content {
    font-size: 20px;
}


</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dpi=300)
library(shiny)
library(ggplot2)
library(forecast)
library(plotly)
library(dplyr)
library(igraph)
library(reshape)
library(spData)
library(leaflet)
library(readr)
library(LaplacesDemon)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(hrbrthemes)
library(gridExtra)
library(cowplot)
library(viridis)
library(gapminder)
library(knitr)
library(kableExtra)
library(DT)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", 
colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  blue = "#1E90FF",
  white = "#FFFFFF"
))
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(MASS) # for truehist function
load("Daily_CMX_means2.Rda")

```
---
## Roadmap


### This class
- Testing assumptions behind residuals
  - Linearity
  - Constant Variance
  - Uncorrelated Residuals
  - Normality
  
---


Let's revisit the assumptions behind linear model:


0. Model is linear in the parameter and with additive error term
1. $E(u_i)=0$
2. $Var(u_i)=\sigma^2$
3. $cov(u_i,u_j)=0$

Additional assumption needed for hypothesis testing and confidence intervals:

5. $u_i \sim N(0,\sigma)$



---
### Assumptions


Visualizing residuals, we can test: 
- Linearity
- Constant Variance (homoskedasticity)
- Uncorrelated errors

---

With vizuale and numerical tests we can analyze:
- Normality

---

### Linearity diagnostic

1. Plot residuals against fitted values of y $\hat{y}$
2. Check if residuals have systematic non-linear pattern

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
set.seed(123)
x <- seq(-5, 5, length.out = 70)
y <- 2 + 2 * x +x^2+ rnorm(70, sd = 3)  # Linear relationship

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
library(ggplot2)
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side

grid.arrange(gg1, gg2, ncol = 2)


```

 
---
### Linearity diagnostic

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
set.seed(127)
x <- seq(-5, 5, length.out = 70)
y <- 2 + 2 * x+ rnorm(70, sd = 3)  # Linear relationship

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
library(ggplot2)
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side

grid.arrange(gg1, gg2, ncol = 2)


```

---
### Constant variance (homoskedasticity)  diagnostic


1. Plot residuals against fitted values of y $\hat{y}$
2. Check if variance changes as $\hat{y}$ changes (heteroskedasticity)


```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data with heteroskedasticity (decreasing variance as x increases)
set.seed(123)
x <- seq(0.1, 5, length.out = 100)
y <- 2 + 3 * x + rnorm(100, sd = 0.9 * (1 - x/5))  # Heteroskedasticity

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side
grid.arrange(gg1, gg2, ncol = 2)

```

---
### Constant variance (homoskedasticity) diagnostic


```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data with heteroskedasticity (decreasing variance as x increases)
set.seed(123)
x <- seq(0.1, 5, length.out = 100)
y <- 2 + 3 * x + rnorm(100, sd = 0.9)  # Heteroskedasticity

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals
gg1 <- ggplot(data, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue", formula = y ~ x) +
  geom_segment(aes(xend = x, yend = predict(linear_model, data), x = x, y = y), color = "red", alpha = 0.2, linetype = "dashed") +
  labs(x = "X", y = "Y") +
  theme_xaringan() +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black")

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

# Combine the two plots side by side
grid.arrange(gg1, gg2, ncol = 2)

```


---
What goes wrong here?

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data for a linear relationship
# Generate sample data for a linear relationship
set.seed(123)
x <- seq(0.1, 5, length.out = 100)  # Avoid x = 0 for division
y <- 2 + 1 / x + rnorm(100, sd = 0.3)  # Relationship like 1/x

# Create a data frame
data <- data.frame(x, y)

# Fit a simple linear regression model in x
linear_model <- lm(y ~ x, data)

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_xaringan()

gg2
```
---

What goes wrong here?

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Generate sample data with changing variance
set.seed(123)
n <- 100
x <- seq(-5, 5, length.out = n)
y <- 2 + 2 * x + rnorm(n, sd = 1 + abs(x))  # Increasing and then decreasing residual variance

# Create a data frame
data <- data.frame(x, y)

# Fit a linear regression model
linear_model <- lm(y ~ x, data)

# Create a scatter plot with the linear regression line and residuals

# Create a residuals vs. fitted values plot for the linear model
fitted_values_linear <- predict(linear_model, data)
residuals_linear <- residuals(linear_model)
gg2 <- ggplot(data.frame(Fitted = fitted_values_linear, Residuals = residuals_linear), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_segment(aes(xend = Fitted, yend = 0, x = Fitted, y = Residuals), color = "blue", alpha = 0.2, linetype = "dashed") +
  labs(x = "Fitted Values", y = "Residuals") +
  theme_minimal()

# Combine the two plots side by side
gg2
```


---
### Correlation of error terms

1. Plot fitted residuals vs their value in previous observation $e_i$ vs $e_{i-1}$
2. So $e_2$ vs $e_1$, $e_5$ vs $e_4$ etc

```{r, warning=FALSE, fig.height=4, out.width='100%'}
# Set seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Generate data with correlated residuals
correlation_values <- c(0.8, 0.6, 0.4, 0.2, 0)  # Decreasing correlation values
scatterplot_list <- list()

# Initialize a list to store autocorrelation coefficients
autocorrelation_coefficients <- list()
ggplot_list <-vector("list", length = 5)
lag=2
# Create scatterplots for residual vs. lagged residuals

plot_data_column = function (lag) {  # Lag from 1 to 5
  # Create correlated residuals with decreasing correlation
  correlation <- correlation_values[lag]
  cov_matrix <- matrix(c(1, correlation, correlation, 1), ncol = 2)
  correlated_residuals <- mvrnorm(n, c(0, 0), cov_matrix)
  
  # Create a data frame for correlated residuals
  data1 <- data.frame(Residual_i = correlated_residuals[, 1])
  data1[[paste0("Residual_i-", lag)]] <- correlated_residuals[, 2]
  
  autocorr_coeff <- cor(data1$Residual_i, data1[[paste0("Residual_i-", lag)]])

  
ggplot(data=data1, aes(x = Residual_i, y = data1[[paste0("Residual_i-", lag)]])) +
    geom_point(color = "blue") +
    labs(x = "Residual i", y = paste0("Residual_i-", lag)) +
    theme_minimal() +
    ggtitle(paste("Lag:", lag)) +
  geom_text(
    x =-2,  # Adjust the x-coordinate as needed
    y = 2.8,  # Adjust the y-coordinate as needed
    label = paste("AC:", round(autocorr_coeff, 2)),
    hjust = 0,  # Horizontal justification (0 for left-aligned)
    vjust = 1,  # Vertical justification (1 for top-aligned)
    color = "red",  # Text color
    size = 4  # Text size
  )+xlim(-3,3)+ylim(-3,3)
}


plot_data_column(1)
```


---
### Correlation of error terms

1. We can also calculate the correlation with other lags
  - Example: $e_i$ vs $e_{i-3}$ 
  - Note down correlation at each lag: $\small \rho(e_i,e_{i-1})$, $\small  \rho(e_i,e_{i-2})$, $\small  \rho(e_i,e_{i-3})$

```{r, warning=FALSE, fig.height=4, out.width='100%'}

# Set seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Generate data with correlated residuals
correlation_values <- c(0.8, 0.6, 0.4, 0.2, 0)  # Decreasing correlation values
scatterplot_list <- list()

# Initialize a list to store autocorrelation coefficients
autocorrelation_coefficients <- list()
ggplot_list <- vector("list", length = 5)

# Create scatterplots for residual vs. lagged residuals
# Generate data with correlated residuals
correlation_values <- c(0.8, 0.6, 0.4, 0.2, 0)  # Decreasing correlation values
scatterplot_list <- list()

# Initialize a list to store autocorrelation coefficients
autocorrelation_coefficients <- list()

autocor = function (lag) {  # Lag from 1 to 5
  # Create correlated residuals with decreasing correlation
  correlation <- correlation_values[lag]
  cov_matrix <- matrix(c(1, correlation, correlation, 1), ncol = 2)
  correlated_residuals <- mvrnorm(n, c(0, 0), cov_matrix)
  
  # Create a data frame for correlated residuals
  data1 <- data.frame(Residual_i = correlated_residuals[, 1])
  data1[[paste0("Residual_i-", lag)]] <- correlated_residuals[, 2]
  
  autocorr_coeff <- cor(data1$Residual_i, data1[[paste0("Residual_i-", lag)]])
}

# Create scatterplots and collect autocorrelation coefficients
autocorrelation_coefficients=lapply(1:5, autocor)

# Create a bar graph of autocorrelation coefficients
bar_data <- data.frame(Lag = 1:5, Autocorrelation = unlist(autocorrelation_coefficients))
bar_plot <- ggplot(bar_data, aes(x = factor(Lag), y = Autocorrelation)) +
  geom_bar(stat = "identity") +
  labs(x = "Lag", y = "Autocorrelation") +
  theme_xaringan()

ggplot_list <- lapply(1:5, plot_data_column)
# Arrange the scatterplots and bar graph side by side
grid.arrange(
  arrangeGrob(grobs = ggplot_list , ncol = 5),
  bar_plot,
  ncol = 1,
  heights = c(3, 3)  # Adjust the height ratio as needed
)
```
---

### Normality

We can start by looking at histograms
- Some are obvious

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
# Function to generate data with specified residual distribution and plot
generate_and_plot_data <- function(residual_dist, df = 100) {
  # Generate data with specified residual distribution
  set.seed(123)
  x <- rnorm(200)
  epsilon <- switch(
    residual_dist,
    "normal" = rnorm(200, sd = 3),
    "exponential" = rexp(200, rate = 1/3),
    "t" = rt(200, df = df)
  )
  y <- 2 + 3 * x + epsilon
  
  # Create a data frame
  data <- data.frame(x, y)
  
  # Fit a linear regression model
  model <- lm(y ~ x, data)
  
  # Create a scatterplot with fitted regression line
  scatterplot <- ggplot(data, aes(x = x, y = y)) +
    geom_point(color = "black") +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(x = "X", y = "Y") +
    theme_xaringan()
  
  # Calculate residuals
  residuals <- resid(model)
  
  # Create a histogram of residuals
  histogram <- ggplot(data.frame(Residuals = residuals), aes(x = Residuals)) +
    geom_histogram(binwidth = 0.5, fill = "green", color = "black", alpha = 0.7) +
    labs(x = "Residuals", y = "Frequency") +
    theme_xaringan()
  
  # Arrange the scatterplot and histogram side by side
  grid.arrange(scatterplot, histogram, ncol = 2)
}

# Create three sets of scatterplots and histograms with different residual distributions
generate_and_plot_data("normal")

```

---
### Normality
We can start by looking at histograms
- Some are obvious


```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
generate_and_plot_data("exponential")

```

---
### Normality
We can start by looking at histograms
- Some are **NOT** obvious
- Then we need a different test



```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
generate_and_plot_data("t", df = 3)
```

---
### Normality


**Q-Q Plot**: Quantlie-Quantile Plot

- Comparing quantiles of our data to what they should be if they come from normal distribution


- Procedure:
  - Standardize residuals

--
  - Sort them

--
  - Check which quantile they represent
    - This is equivalent to checking cumulative probability: $CP_i=\frac{Index}{n+1}$
    - Where index is the number of observation if we sort if from smallest to  largest 
        - 1 is the smallest, 2 is the second smallest, ..., n is the largest
    - In other words, what share of data is smaller than this observation

--
  - Compare them to quantiles from standard normal

--
  - If our distribution is normal, the quantiles should be similar

---
### Q-Q plot

```{r, echo=FALSE, results='asis'}
# Create a data frame for the table
# Set the seed for reproducibility
set.seed(123)

# Generate normally distributed residuals with standard deviation = 3
residuals <- round(rnorm(100, sd = 3),3)

# Standardize the residuals
standardized_residuals <- round(scale(residuals),3)

# Create a data frame with the residuals and standardized residuals
data_df <- data.frame(Residuals = residuals, Standardized_Residuals = standardized_residuals)

# Calculate the order (rank) of standardized residuals from smallest to largest
data_df$Index <- rank(data_df$Standardized_Residuals)
data_df$Cumulative_Probability <- round(data_df$Index / (nrow(data_df) + 1),3)
data_df$Quantile_Normal <- round(qnorm(data_df$Cumulative_Probability),3)


# Display the data in a table
datatable(data_df,
          fillContainer = FALSE,
          options = list(
            pageLength = 10,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE
)
```
---
### Q-Q plot

- Next, we plot the sample quantiles vs the standard normal quantiles
- If they are similar, they should be on the straight line

### Q-Q plot
.pull-left[
```{r,include=FALSE,  warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
# Set the seed for reproducibility
# Create a histogram plot with ggplot2
histogram_plot <- ggplot(data_df, aes(x = scale(Residuals))) +
  geom_histogram(binwidth = 0.4, aes(y = after_stat(density)), fill = "green", color = "black", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), color = "blue") +  # Add standard normal density
  labs(x = "Standardized Residuals", y = "Density") +
  theme_minimal()+
  xlim(-3,3)

# Create a QQ plot with ggplot2

# Convert both plots to plotly objects
histogram_plot <- ggplotly(histogram_plot, tooltip = "y",
        width = 400,   # Adjust the width according to your preference
        height = 350)
```

```{r, warning=FALSE, fig.height=3.5, out.width='100%', message=FALSE}
histogram_plot

```
]
.pull-right[
```{r, include=FALSE, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
qq_plot <- ggplot(data_df) +
  geom_qq(aes(sample = scale(Residuals))) +
  geom_abline(color = "red") +
  coord_fixed() +
  theme_minimal() +
  xlim(-3, 3) +
  ylim(-3, 3)  # Set y-axis limits to show cumulative probability




qq_plot=ggplotly(qq_plot,
                    width = 400,   # Adjust the width according to your preference
                    height = 350)

data_df$Index <- rank(data_df$Residuals)
data_df$Cumulative_Probability <- round(data_df$Index / (nrow(data_df) + 1),3)
data_df <- data_df[order(data_df$Residuals), ]
qq_plot$x$data[[1]]$text <- paste("Quantile:", data_df$Cumulative_Probability, "\nTheoretical:", data_df$Quantile_Normal,"\nSample: ", data_df$Standardized_Residuals)

```

```{r, warning=FALSE, fig.height=3.5, out.width='100%', message=FALSE}
qq_plot
```
]



---
### Q-Q plot
.pull-left[
```{r,include=FALSE,  warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
# Set the seed for reproducibility
library(plotly)

# Generate chi-square distributed residuals with 3 degrees of freedom
residuals <- rchisq(100, df = 3)

# Create a data frame with the residuals
data_df <- data.frame(Residuals = scale(residuals))
data_df$Standardized_Residuals=data_df$Residuals
data_df$Index <- rank(data_df$Standardized_Residuals)
data_df$Cumulative_Probability <- round(data_df$Index / (nrow(data_df) + 1),3)
data_df$Quantile_Normal <- round(qnorm(data_df$Cumulative_Probability),3)


# Create a histogram plot with ggplot2
histogram_plot <- ggplot(data_df, aes(x = scale(Residuals))) +
  geom_histogram(binwidth = 0.4, aes(y = after_stat(density)), fill = "green", color = "black", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1), color = "blue") +  # Add standard normal density
  labs(x = "Standardized Residuals", y = "Density") +
  theme_minimal()+
  xlim(-3,3)

# Create a QQ plot with ggplot2

# Convert both plots to plotly objects
histogram_plot <- ggplotly(histogram_plot, tooltip = "y",
        width = 400,   # Adjust the width according to your preference
        height = 350)
```

```{r, warning=FALSE, fig.height=3.5, out.width='100%', message=FALSE}
histogram_plot

```
]
.pull-right[
```{r, include=FALSE, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}

qq_plot <- ggplot(data_df) +
  geom_qq(aes(sample = Residuals)) +
  geom_abline(color = "red") +
  coord_fixed() +
  theme_minimal() +
  xlim(-3, 3) +
  ylim(-3, 3)  # Set y-axis limits to show cumulative probability




qq_plot=ggplotly(qq_plot,
                    width = 400,   # Adjust the width according to your preference
                    height = 350)

data_df$Index <- rank(data_df$Residuals)
data_df$Cumulative_Probability <- round(data_df$Index / (nrow(data_df) + 1),3)
data_df <- data_df[order(data_df$Residuals), ]
qq_plot$x$data[[1]]$text <- paste("Quantile:", data_df$Cumulative_Probability, "\nTheoretical:", data_df$Quantile_Normal,"\nSample: ", round(data_df$Standardized_Residuals,3))

```

```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}
qq_plot
```
]


---
### Q-Q plot

```{r, warning=FALSE, fig.height=3, out.width='100%', message=FALSE}


# Define the number of data points
num_data_points <- 100

# Generate uniform random residuals between -1 and 1
uniform_residuals <- runif(num_data_points, min = -2, max = 2)

# Create independent variable (predictor)
x <- rnorm(num_data_points)

# Create dependent variable (response) with a linear relationship and added residuals
y <- -2 * x + 3 + uniform_residuals

# Create a data frame with predictor and response variables
data_df <- data.frame(x, y)

residuals=uniform_residuals

# Create a fitted regression plot
regression_plot <- ggplot(data_df, aes(x = x, y = y)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "X", y = "Y") +
  theme_minimal()

# Create a histogram of standardized residuals
histogram_plot <- ggplot(data_df, aes(x = scale(residuals))) +
  geom_histogram(binwidth = 0.4, aes(y = after_stat(density)), fill = "green", color = "black", alpha = 0.7) +
 stat_function(fun = dnorm, args = list(mean = 0, sd = 1), color = "blue") +
  labs(x = "Standardized Residuals", y = "Density") +
  theme_minimal()+
  xlim(-3,3)

# Create a QQ plot
qq_plot <- ggplot(data_df) +
  geom_qq(aes(sample = scale(residuals))) +
  geom_abline(color = "red") +
  coord_fixed() +
  theme_minimal()+
  xlim(-3,3)+
  ylim(-3,3)

# Arrange the three plots side by side
grid.arrange(regression_plot, histogram_plot, qq_plot, ncol = 3)

```


---
### Jarque-Bera Test

- We can also use the **Jarque-Bera Test** to see whether sample comes from a normal distribution

**Intuition**
- Skeweness of normal distribution is 0 (it's symmetric)

$$S = \frac{\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^3}{\left(\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{\frac{3}{2}}}$$

---


```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}

# Set the seed for reproducibility
set.seed(123)

# Generate data with skewness = 0 (normally distributed)
data_0_skew <- rnorm(1000)

# Generate data with positive skewness
data_pos_skew <- c(rnorm(800), rnorm(200, mean = 3, sd = 1))

# Create a data frame for each dataset
df_0_skew <- data.frame(Data = data_0_skew, Group = "Skewness = 0")
df_pos_skew <- data.frame(Data = data_pos_skew, Group = "Positive Skewness")

# Combine the data frames
combined_df <- rbind(df_0_skew, df_pos_skew)

# Create a density plot for each dataset without legends
density_plot_0_skew <- ggplot(df_0_skew, aes(x = Data)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot (Skewness = 0)") +
  theme_minimal() +
  theme(legend.position = "none")

density_plot_pos_skew <- ggplot(df_pos_skew, aes(x = Data)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot (Positive Skewness)") +
  theme_minimal() +
  theme(legend.position = "none")

# Determine the common x-axis limits
x_limits <- range(combined_df$Data)

# Arrange the density plots side by side with equal x scales
grid.arrange(
  density_plot_0_skew + xlim(-6,6),
  density_plot_pos_skew + xlim(-6,6),
  ncol = 2
)
```

---
### Jarque-Bera Test

- We can also use the **Jarque-Bera Test** to see whether sample comes from a normal distribution

**Intuition**
- Skeweness of normal distribution is 0 (it's symmetric)

$$S = \frac{\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^3}{\left(\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{\frac{3}{2}}}$$


- Kurtosis of normal distribution is 3 
- Excess kurtosis is 0

$$EK = \frac{\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^4}{\left(\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{2}} - 3$$


---


```{r, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}

# Set the seed for reproducibility
set.seed(123)

generate_data <- function(n, kurtosis) {
  if (kurtosis == 3) {
    # Generate normally distributed data
    data <- rnorm(n)
  } else if (kurtosis < 3) {
    # Generate data with specified kurtosis using the PDF f(x) = 6x(1 - x)
    mu <- 0
    b <- 0.6
    data <- rlaplace(n, mu, b)
  } else {
    # Generate data with specified kurtosis using the PDF f(x) = 1/(πx(1 - x))√
    u <- runif(n)
    data <- sqrt(6 * (kurtosis - 3)) * tan(pi * (u - 0.5))
  }
  return(data)
}

# Generate data with different kurtosis values
n <- 10000
data_kurtosis_less_3 <- generate_data(n, kurtosis = 2.3)
data_kurtosis_3 <- generate_data(n, kurtosis = 3)
data_kurtosis_greater_3 <- generate_data(n, kurtosis = 5)

# Create data frames for each dataset
df_kurtosis_less_3 <- data.frame(Data = data_kurtosis_less_3, Group = "Kurtosis < 3")
df_kurtosis_3 <- data.frame(Data = data_kurtosis_3, Group = "Kurtosis = 3")
df_kurtosis_greater_3 <- data.frame(Data = data_kurtosis_greater_3, Group = "Kurtosis > 3")

# Combine the data frames
combined_df <- rbind(df_kurtosis_less_3, df_kurtosis_3, df_kurtosis_greater_3)

# Create density plots for each dataset without legends
density_plot_kurtosis_less_3 <- ggplot(df_kurtosis_less_3, aes(x = Data)) +
  geom_density(alpha = 0.5) +
  labs(title = "Kurtosis < 3") +
  theme_minimal() +
  theme(legend.position = "none")+
  xlim(-10,10)+
  geom_density(data = data.frame(Data = rnorm(10000, 0, 1)), aes(x = Data), color = "blue")


density_plot_kurtosis_3 <- ggplot(df_kurtosis_3, aes(x = Data)) +
  geom_density(alpha = 0.5) +
  labs(title = "Kurtosis = 3") +
  theme_minimal() +
  theme(legend.position = "none")+
  xlim(-10,10)

density_plot_kurtosis_greater_3 <- ggplot(df_kurtosis_greater_3, aes(x = Data)) +
  geom_density(alpha = 0.5) +
  labs(title = "Kurtosis > 3") +
  theme_minimal() +
  theme(legend.position = "none")+
  xlim(-10,10)+
  geom_density(data = data.frame(Data = rnorm(10000, 0, 1)), aes(x = Data), color = "blue")


# Determine the common x-axis limits
x_limits <- range(combined_df$Data)

# Arrange the density plots side by side with equal x scales
grid.arrange(
  density_plot_kurtosis_less_3 + xlim(-10,10)+ ylim(0,0.8),
  density_plot_kurtosis_3 + xlim(-10,10)+ylim(0,0.8),
  density_plot_kurtosis_greater_3 + xlim(-10,10)+ylim(0,0.8),
  ncol = 3
)
```


---
### Jarque-Bera Test

- We can also use the **Jarque-Bera Test** to see whether sample comes from a normal distribution

**Intuition**
- Skeweness of normal distribution is 0 (it's symmetric)

$$S = \frac{\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^3}{\left(\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{\frac{3}{2}}}$$


- Kurtosis of normal distribution is 3 
- Excess kurtosis is 0

$$EK = \frac{\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^4}{\left(\frac{1}{n} \sum_{i=1}^{n}(x_i - \bar{x})^2\right)^{2}} - 3$$


- We will test, whether our sample has (more or less)
  - Skewenss of 0
  - Kurtosis of 3

---
### Jarque-Bera Test


The test statistic for our test is:

$$JB = \frac{n}{6} \left(S^2 + \frac{EK^2}{4}\right)$$

- Its value will be high if:
  - Skeweness deviates significantly from 0
  - Kurtosis deviates significantly from 3

--

- If the data really comes from normal (that's our null hypothesis), then:

$$JB \sim  \chi_2$$
- It follows the Chi-squared distribution with 2 degrees of freedom. 

---
### Jarque-Bera Test

So in our usual testing setting:

- $H_0: e_i$ comes from normal (JB is small)
- $H_A: e_i$ Does not come from normal (JB is large)

- It's a one sided test, so we reject at $\alpha$ if 

$$JB_{test}>\chi_{1-\alpha,2}$$
---
### Jarque-Bera Test

```{r, warning=FALSE, fig.height=2.5, out.width='100%', message=FALSE}

# Generate a sample of 100 values from a normal distribution with mean 0 and standard deviation 1
sample_data <- rnorm(200, mean = 10, sd = 2)

# Create a histogram using ggplot2
ggplot(data = data.frame(x = sample_data), aes(x = x)) +
  geom_histogram(binwidth = 0.3, color="white") +
  labs(x = "Values",
       y = "Frequency")+
  theme_xaringan()


```

```{r, echo=TRUE, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}

library(tseries)
jarque.bera.test(sample_data)
```

---
### Jarque-Bera Test

```{r, warning=FALSE, fig.height=2.5, out.width='100%', message=FALSE}

# Generate a sample of 100 values from a normal distribution with mean 0 and standard deviation 1
sample_data_exponential <- rexp(200, rate = 1)
sample_data=scale(sample_data_exponential)

# Create a histogram using ggplot2
ggplot(data = data.frame(x = sample_data), aes(x = x)) +
  geom_histogram(binwidth = 0.3, color="white") +
  labs(x = "Values",
       y = "Frequency")+
  theme_xaringan()


```

```{r, echo=TRUE, warning=FALSE, fig.height=4, out.width='100%', message=FALSE}

library(tseries)
jarque.bera.test(sample_data)
```

---