---
title: "Multiple Linear Regression"
author: "Business Forecasting"
output:
  learnr::tutorial:
    progressive: false
    allow_skip: true
runtime: shiny_prerendered
tutorial:
  id: "itam-bf-C5-mlr"
  version: 2.0
description: "Learn to estimate, interpret, and test multiple linear regression models"
---


```{r setup, include=FALSE}
for (pkg in c("learnr","shiny","tidyverse","plotly","tseries","car","probstats4econ","wooldridge")) {
  if (!requireNamespace(pkg, quietly = TRUE))
    install.packages(pkg, repos = "https://cloud.r-project.org")
  library(pkg, character.only = TRUE)
}

  # Fixed data generation
  set.seed(123) # For reproducibility
  x <- seq(-10, 10, by = 1)
  e <- rnorm(length(x), mean = 0, sd = 5)
  y_actual <- 4 + 1.65 * x + e

  data <- data.frame(x, y_actual)


```

## 1.1 Matrix OLS: Predicting Affairs

| | |
|---|---|
| **Data** | `affairs` (wooldridge) |
| **Y** | `naffairs` — number of affairs |
| **X** | `age` — age (years); `relig` — religiosity (1-5) |

We'll work with data from a survey on extra-marital affairs in the US. Here is the data:

- **id**: identifier
- **male**: =1 if male
- **age**: age in years
- **yrsmarr**: years married
- **kids**: =1 if have kids
- **relig**: religiosity (5 = very religious, 4 = somewhat, 3 = slightly, 2 = not at all, 1 = anti-religious)
- **educ**: years of schooling
- **ratemarr**: marital happiness (5 = very happy, 4 = happier than average, 3 = average, 2 = somewhat unhappy, 1 = very unhappy)
- **naffairs**: number of affairs within last year
- **affair**: =1 if had at least one affair


Let's predict the number of affairs. Let's pick two variables and find the coefficients:

\[
y_i=\beta_0+\beta_1x_{i3}+\beta_2x_{i2}+\epsilon_i
\]

Let's find the coefficients using the OLS formula:
$$\beta=(X'X)^{-1}X'y$$


```{r e1, exercise=TRUE}
# Build X matrix with cbind(1, ...), compute solve(t(X) %*% X) %*% (t(X) %*% y)

```



```{r e2, exercise=TRUE}
# Use lm(y ~ x1 + x2, data = dataset) and summary()

```


Find predicted values

```{r e3, exercise=TRUE, exercise.setup="e2"}
# yhat = X %*% beta_hat, or use model$fitted.values

```



Find residuals

```{r e4, exercise=TRUE, exercise.setup="e2"}
# residuals = y - yhat, or use model$residuals

```

Make a prediction for particular values of x vector using beta vector from the first derivation.

```{r e5, exercise=TRUE, exercise.setup="e1"}
# x_new = c(1, value1, value2, ...); y_pred = x_new %*% beta_hat

```

---

## 1.2 Your Turn: Real Estate Pricing Model

| | |
|---|---|
| **Data** | `hprice1` (wooldridge) |
| **Y** | `price` — house price (thousands $) |
| **X** | `bdrms` — bedrooms; `lotsize` — lot (sq ft); `sqrft` — house size (sq ft) |

**Business Context:** A real estate agency wants to build a pricing model. Use `hprice1` from wooldridge: `price` (in $1000s), `bdrms` (bedrooms), `lotsize` (lot size in sqft), `sqrft` (house size sqft).

**Your Decision:** Which features drive house prices the most?

Use the matrix formula $\beta=(X'X)^{-1}X'y$ to find the coefficients. Then verify your result with `lm()`.

```{r biz_c5_1, exercise=TRUE}
# Your code here

```

## 2.1 Omitted Variables: Wages, Education & IQ

| | |
|---|---|
| **Data** | `card` (wooldridge) |
| **Y** | `wage` — weekly wage (cents) |
| **X** | `educ` — years of education; `IQ` — IQ score |

Using dataset "card", regress wage (weekly, in cents) on educ (education in years). Then add IQ. How do the coefficient on education changes?



```{r e6, exercise=TRUE}
# Run lm(wage ~ educ) then lm(wage ~ educ + IQ) and compare coefficients with summary()

```

---

## 2.2 Your Turn: CEO Pay Beyond ROE

| | |
|---|---|
| **Data** | `ceosal1` (wooldridge) |
| **Y** | `salary` — CEO salary (thousands $) |
| **X** | `roe` — return on equity (%); `sales` — firm sales (millions $) |

**Business Context:** A board of directors found that ROE explains CEO salary. But does adding firm sales improve the model? Use `ceosal1`: `salary`, `roe`, `sales`.

**Your Decision:** Is firm size a confounding factor in the ROE-salary relationship?

Run `salary ~ roe`, then `salary ~ roe + sales`. Compare the coefficient on `roe`.

```{r biz_c5_2, exercise=TRUE}
# Your code here

```

## 3.1 Variance-Covariance: Affairs Model

| | |
|---|---|
| **Data** | `affairs` (wooldridge) |
| **Y** | `naffairs` — number of affairs |
| **X** | `age`; `relig` |

We'll work with data from a survey on extra-marital affairs in the US. Regress nafffairs on age and relig. Find the variance covaraince matrix.


```{r e7, exercise=TRUE}
# Use vcov(model), or compute sigma2 * solve(t(X) %*% X)

```

---

## 3.2 Your Turn: CEO Salary VCOV

| | |
|---|---|
| **Data** | `ceosal1` (wooldridge) |
| **Y** | `salary` — CEO salary (thousands $) |
| **X** | `roe` — return on equity (%); `sales` — firm sales (millions $) |

**Business Context:** The board needs standard errors for their CEO compensation model to know how reliable each estimate is. Use `ceosal1`: `salary`, `roe`, `sales`.

Compute the variance-covariance matrix for `salary ~ roe + sales` using the matrix formula. Verify with `vcov()`.

```{r biz_c5_3, exercise=TRUE}
# Your code here

```

## 4.1 Categorical Variables: Wages by Marital Status

| | |
|---|---|
| **Data** | `cps` (probstats4econ) |
| **Y** | `wagehr` — hourly wage ($) |
| **X** | `marstatus` — Married/Divorced/Widowed/Never married |

How does the marital status impacts earning?

Let's use cps data (survey  in the US). It provides information on both hourly wage (wagehr) and marital status (marstatus).  Possible options are: Married, Divorced, Widowed, Never married.


```{r e8, exercise=TRUE}
# Use lm(y ~ factor(x), data = dataset)

```

---

## 4.2 Your Turn: Gender Pay Gap in Hourly Wages

| | |
|---|---|
| **Data** | `wage1` (wooldridge) |
| **Y** | `wage` — hourly wage ($) |
| **X** | `female` — 1 if female |

**Business Context:** An HR department needs to quantify the gender wage gap. Use `wage1`: `wage` (hourly $) and `female` (1 if female).

**Your Decision:** What is the average pay difference? Is it significant?

Regress `wage` on `female`. Interpret $\beta_0$ (average male wage) and $\beta_1$ (the gap).

```{r biz_c5_4, exercise=TRUE}
# Your code here

```

## 5.1 Interaction: House Size x Year Built

| | |
|---|---|
| **Data** | `houseprices` (probstats4econ) |
| **Y** | `saleprice` — sale price ($) |
| **X** | `lotarea` — lot size (sq ft); `yearbuilt` — year built |

Let's look at how value of the size of the house changes with its age. This data reports houses sold between 2006 and 2010. It has three variables of interest:

1. salesprice - sale price
2. lotarea - size of the lot
3. yearbuilt - year the house was built



```{r e9, exercise=TRUE}
# Use lm(y ~ x1 * x2, data = dataset)

```

---

## 5.2 Your Turn: Experience x Tenure

| | |
|---|---|
| **Data** | `wage1` (wooldridge) |
| **Y** | `wage` — hourly wage ($) |
| **X** | `exper` — experience (years); `tenure` — tenure (years) |

**Business Context:** A compensation analyst suspects that the value of general work experience depends on how long the person has been at the company. Use `wage1`: `wage`, `exper`, `tenure`.

**Your Decision:** Should we design different pay scales for new vs. long-tenured employees?

Run `wage ~ exper * tenure`. Interpret the interaction term.

```{r biz_c5_5, exercise=TRUE}
# Your code here

```

---

## 5.3 Interaction: Education by Race

| | |
|---|---|
| **Data** | `cps` (probstats4econ) |
| **Y** | `wagehr` — hourly wage ($) |
| **X** | `educ` — years of education; `race` — Black/White/Other |

How do return to education vary across racial groups?

Let's use cps data (survey  in the US). It provides information on both hourly wage (wagehr) and racial group (race: Black, White, Other)  and education in years (educ).

```{r 10, exercise=TRUE}
# Use lm(y ~ x1 * x2, data = dataset) for interaction between education and race

```

---

## 5.4 Your Turn: Gender Gap by Education Level

| | |
|---|---|
| **Data** | `wage1` (wooldridge) |
| **Y** | `wage` — hourly wage ($) |
| **X** | `educ` — education (years); `female` — 1 if female |

**Business Context:** An equal-pay advocate claims the gender gap widens with education. Use `wage1`: `wage`, `educ`, `female`.

**Your Decision:** Should pay equity policies focus on higher-educated roles?

Run `wage ~ educ * female`. Interpret the interaction term.

```{r biz_c5_6, exercise=TRUE}
# Your code here

```

## 6.1 R-squared: Student Exam Performance

| | |
|---|---|
| **Data** | `attend` (wooldridge) |
| **Y** | `final` — final exam score (0-40) |
| **X** | `attend`, `ACT`, `termGPA`, `priGPA`, etc. |

Data "attend" reports final score of students of microeconomics from a US universities. It has following variables:

- attend: classes attended out of 32
- termGPA: GPA for term (GPA: grade point average)
- priGPA: cumulative GPA prior to term
- ACT: ACT score (standarized end of high school exam)
- final: final exam score
- atndrte: percent classes attended
- hwrte: percent homework turned in
- frosh: =1 if freshman
- soph: =1 if sophomore
- missed: number of classes missed


Outcome is final exam score. It goes from 0 to 40. (that's what we want to predict). Keep adding regressors and try to find the highest adjusted R square possible. (ex: attend, ACT, termGPA, priGPA)

```{r 11, exercise=TRUE}
# Compare summary(model)$adj.r.squared across models with different regressors

```

---

## 6.2 Your Turn: Baseball Player Salaries

| | |
|---|---|
| **Data** | `mlb1` (wooldridge) |
| **Y** | `salary` — annual salary (thousands $) |
| **X** | `years`, `games`, `bavg`, `hrunsyr`, `rbisyr` |

**Business Context:** A sports analytics team is building a player valuation model. Use `mlb1`: `salary` (annual in $1000s), `years` (in league), `games` (career games), `bavg` (batting avg), `hrunsyr` (home runs/yr), `rbisyr` (RBIs/yr).

**Your Decision:** Which stats should scouts prioritize when negotiating contracts?

Add predictors one by one and track adjusted $R^2$. Find the best model.

```{r biz_c5_7, exercise=TRUE}
# Your code here

```

## 7.1 Hypothesis Test: Alcohol & Employment

| | |
|---|---|
| **Data** | `alcohol` (wooldridge) |
| **Y** | `employ` — employed (0/1) |
| **X** | `abuse` — alcohol abuse (0/1); `educ` — education (years) |

In the dataset "alcohol" we find information whether someone is employed (employ), whether they abuse alcohol (abuse), and years of education (educ).
Test if abusing alcohol affects your employment status when controlling for education. Compare to a model that doesn't include education as control.

```{r 13f, exercise=TRUE}
# Check the t-statistic and p-value from summary()
# What is the distribution of the test statistic? What is the number of degrees of freedom?

```

---

## 7.2 Your Turn: Does Sleep Protect Wages?

| | |
|---|---|
| **Data** | `sleep75` (wooldridge) |
| **Y** | `hrwage` — hourly wage ($) |
| **X** | `slpnaps` — total sleep (hours/week); `educ` — education (years); `age` |

**Business Context:** A wellness program manager asks: does getting more sleep improve hourly wages even after accounting for education and age? Use `sleep75`: `hrwage`, `slpnaps`, `educ`, `age`.

**Your Decision:** Should we invest in employee wellness and sleep programs?

Test the significance of `slpnaps` controlling for `educ` and `age`.

```{r biz_c5_8, exercise=TRUE}
# Your code here

```

---

## 7.3 Interaction Test: Education & City Size

| | |
|---|---|
| **Data** | `beauty` (wooldridge) |
| **Y** | `wage` — hourly wage ($) |
| **X** | `educ` — education (years); `bigcity` — 1 if big city |

Does education plays larger role if living in a big city?

The dataset "beauty" has information on wage (wage), years of education (educ), and whether someone lives in a big city (bigcity). Test if return to education is higher when you live in big city.

```{r 13, exercise=TRUE}
# Use lm(y ~ x1 * x2, data = dataset) and check the t-statistic on the interaction term
# What is the distribution of the test statistic? What is the number of degrees of freedom?

```

---

## 7.4 Your Turn: City Size & Returns to Education

| | |
|---|---|
| **Data** | `beauty` (wooldridge) |
| **Y** | `wage` — hourly wage ($) |
| **X** | `educ` — education (years); `bigcity` — 1 if big city |

**Business Context:** A workforce study asks if workers in big cities get a higher return to education. Use `beauty`: `wage`, `educ`, `bigcity`.

**Your Decision:** Is the education premium different in big vs small cities?

Run `wage ~ educ * bigcity` and test the interaction term.

```{r biz_c5_9, exercise=TRUE}
# Your code here

```

## 8.1 CI for Coefficients: Sleep & Wages

| | |
|---|---|
| **Data** | `sleep75` (wooldridge) |
| **Y** | `slpnaps` — total sleep (hours/week) |
| **X** | `hrwage` — hourly wage ($); `age`; `marr` — married (0/1) |

The dataset sleep75 contains informatiomn on total hours of sleep (slpnaps), age, being married (marr), and hourly wage (hrwage).
Run a regression explaining sleep time using all these variables. What is 95% confidence interval on the impact of more dollar of hourly wage? What about being married?


```{r 14, exercise=TRUE}
# Use confint(model, level = 0.95)
# What is the distribution? What is the number of degrees of freedom?

```

---

## 8.2 Your Turn: Value of Years in the League

| | |
|---|---|
| **Data** | `mlb1` (wooldridge) |
| **Y** | `salary` — annual salary (thousands $) |
| **X** | `years` — years in league; `bavg` — batting avg; `hrunsyr` — home runs/yr |

**Business Context:** A sports analytics team is designing seniority-based pay adjustments. Use `mlb1`: `salary`, `years`, `bavg`, `hrunsyr`.

**Your Decision:** What range of salary increase per additional year in the league should we budget?

Compute the 95% confidence interval for the `years` coefficient in `salary ~ years + bavg + hrunsyr`.

```{r biz_c5_10, exercise=TRUE}
# Your code here

```

---

## 8.3 CI for Mean Response: Student Scores

| | |
|---|---|
| **Data** | `econmath` (wooldridge) |
| **Y** | `score` — final score |
| **X** | `age`, `study`, `act`, `fathcoll`, `mothcoll`, `male` |

The dataset econmath contains data from a course in economics from Michigan State University. It reports the following variables.

- score: final score
- age: age in years
- study: hours studying per week
- act: score on final high school exam
- fathcoll: 1 if father has college degree
- mothcoll: 1 if mother has college degree
- male: 1 if male

Let's find the confidence interval for the mean score for students who are 22 years old, study 10 hours per week, had 25 on final high school exam, only mother has a college degree, and is male

#### Step by step method:

1. Find the prediction
2. Write down the vector of predictors
3. Find the inverse of X'X
4. Find the standard error
5. Compute the standard errors of the prediction
6. Find critical value
7. Put together the interval

```{r 15m1, exercise=TRUE}
# Step by step: fit lm(), define x0 vector, compute se = sigma * sqrt(t(x0) %*% solve(t(X) %*% X) %*% x0)
# Then build interval: yhat +/- t_crit * se

```

#### Quick method:

```{r 15m2, exercise=TRUE}
# Use predict(model, newdata = ..., interval = "confidence", level = 0.95, se.fit = TRUE)

```

---

## 8.4 Your Turn: Expected Price for 3-Bed House

| | |
|---|---|
| **Data** | `hprice1` (wooldridge) |
| **Y** | `price` — house price (thousands $) |
| **X** | `bdrms`, `lotsize`, `sqrft` |

**Business Context:** The real estate agency wants to know the average market price for a 3-bed, 2000 sqft house on an 8000 sqft lot. Use `hprice1`.

**Your Decision:** What average price should we communicate to the seller?

Use `predict(..., interval = "confidence")` with `new_data = data.frame(bdrms = 3, lotsize = 8000, sqrft = 2000)`.

```{r biz_c5_11, exercise=TRUE}
# Your code here

```

---

## 8.5 PI for New Observation: Student Scores

| | |
|---|---|
| **Data** | `econmath` (wooldridge) |
| **Y** | `score` — final score |
| **X** | `age`, `study`, `act`, `fathcoll`, `mothcoll`, `male` |

Let's use the same setting as before, but calculate the confidence interval for a new observation.

#### Step by step method:

1. Find the prediction
2. Write down the vector of predictors
3. Find the inverse of X'X
4. Find the standard error
5. Compute the standard errors of the prediction
6. Find critical value
7. Put together the interval

```{r 16m1, exercise=TRUE}
# Step by step: fit lm(), define x0 vector, compute se = sigma * sqrt(1 + t(x0) %*% solve(t(X) %*% X) %*% x0)
# Then build interval: yhat +/- t_crit * se

```

#### Quick method:

```{r 16m2, exercise=TRUE}
# Use predict(model, newdata = ..., interval = "prediction", level = 0.95, se.fit = TRUE)

```

---

## 8.6 Your Turn: Price This Specific House

| | |
|---|---|
| **Data** | `hprice1` (wooldridge) |
| **Y** | `price` — house price (thousands $) |
| **X** | `bdrms`, `lotsize`, `sqrft` |

**Business Context:** Same house as before (3-bed, 2000 sqft, 8000 sqft lot), but now we need a range for what THIS specific house will sell for.

**Your Decision:** What listing price range should we recommend?

Use `predict(..., interval = "prediction")`. Compare the width of the prediction interval to the confidence interval from above.

```{r biz_c5_12, exercise=TRUE}
# Your code here

```

## 9.1 ANOVA: Fast Food Prices & Demographics

| | |
|---|---|
| **Data** | `discrim` (wooldridge) |
| **Y** | `pfries` — price of fries ($) |
| **X** | `prpblck`, `prppov`, `income` |

The dataset discrim contains information of prices of fast food items from the US. Each observation is fast food restaurant.

Suppose you regress price of price of fries (pfries) on location characteristics: proportion black in zipcode (prpblck), proportion living in poverty (prppov), median family income (income). Does this regression helps at all to explain the price?

#### Step 1: Start with running the regression and looking at the results

```{r 17m1, exercise=TRUE}
# Use lm(y ~ x1 + x2 + x3, data = dataset) and summary()

```


#### Step 2: Run the ANOVA test.

```{r 17m2, exercise=TRUE, exercise.setup="17m1"}
# Use anova(model)

```


#### Step 3: What is the SSE, SST and SSR? What are their respective degrees of freedom?

```{r 17m3, exercise=TRUE, exercise.setup="17m2"}
# SSR = sum of Sq from anova, SSE = residual Sum Sq, SST = SSR + SSE
# Degrees of freedom: df_R = k, df_E = n - k - 1

```


#### Step 4: What is the F test result? What is the p-value?

```{r 17m4, exercise=TRUE, exercise.setup="17m3"}
# F = (SSR/k) / (SSE/(n-k-1)), p-value from pf(F, df1, df2, lower.tail = FALSE)

```

---

## 9.2 Your Turn: Do Player Stats Explain Salary?

| | |
|---|---|
| **Data** | `mlb1` (wooldridge) |
| **Y** | `salary` — annual salary (thousands $) |
| **X** | `years`, `bavg`, `hrunsyr`, `rbisyr` |

**Business Context:** The team owner asks: do on-field statistics actually explain salary variation, or is it mostly seniority? Use `mlb1`.

**Your Decision:** Should contract negotiations focus on stats or years of service?

Run `anova(lm(salary ~ years + bavg + hrunsyr + rbisyr, data = mlb1))` and interpret the F-test. Compute SST, SSR, SSE.

```{r biz_c5_13, exercise=TRUE}
# Your code here

```

## 10.1 Coefficient Equality: Spouse Education & Income

| | |
|---|---|
| **Data** | `married` (probstats4econ) |
| **Y** | `famincome` — family income ($) |
| **X** | `educ_w` — wife's education; `educ_h` — husband's education |

Dataset married provides information on married couples in the US. It reports total family income and characteristics of both spouses: age (age_w, age_h), years of education (educ_w, educ_h), and body mass index (bmi_w, bmi_h).

Test if the impact of wife education is the same as the impact of husband education on family income.

#### Step 1: Run the regression

```{r 18m1, exercise=TRUE}
# Use lm(famincome ~ educ_w + educ_h, data = married) and summary()

```


#### Step 2: Find the test statistic

```{r 18m2, exercise=TRUE, exercise.setup="18m1"}
# t = (b1 - b2) / sqrt(Var(b1) + Var(b2) - 2*Cov(b1,b2)); use vcov(model)

```

#### Step 3: Find the p-value

```{r 18m3, exercise=TRUE, exercise.setup="18m2"}
# p-value = 2 * pt(-abs(t_stat), df = n - k - 1)

```


#### Alternative test.

Now try again but with bmi of wife (bmi_w) vs bmi of husband (bmi_h)


```{r 18m5, exercise=TRUE}
# Use lm(famincome ~ bmi_w + bmi_h, data = married) and test equality of coefficients

```
