---
title: "LearnR Tutorial: Estimators"
author: "Business Forecasting"
output: learnr::tutorial
runtime: shiny_prerendered
---
 

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(tidyverse)

if (!requireNamespace("car", quietly = TRUE)) {
  install.packages("car", repos = "https://cloud.r-project.org")
}

if (!requireNamespace("plotly", quietly = TRUE)) {
  install.packages("plotly", repos = "https://cloud.r-project.org")
}


if (!requireNamespace("tseries", quietly = TRUE)) {
  install.packages("tseries", repos = "https://cloud.r-project.org")
}

library(plotly)
library(car)
library(tseries)

if (!requireNamespace("probstats4econ", quietly = TRUE)) {
  install.packages("probstats4econ", repos = "https://cloud.r-project.org")
}
library(probstats4econ)


if (!requireNamespace("wooldridge", quietly = TRUE)) {
  install.packages("wooldridge", repos = "https://cloud.r-project.org")
}
library(wooldridge)

  
  # Fixed data generation
  set.seed(123) # For reproducibility
  x <- seq(-10, 10, by = 1)
  e <- rnorm(length(x), mean = 0, sd = 5)
  y_actual <- 4 + 1.65 * x + e
  
  data <- data.frame(x, y_actual)
  

```

---

## Deriving Coefficients

We’ll work with data from a survey on extra-marital affairs in the US. Here is the data:

- **id**: identifier  
- **male**: =1 if male  
- **age**: age in years  
- **yrsmarr**: years married  
- **kids**: =1 if have kids  
- **relig**: religiosity (5 = very religious, 4 = somewhat, 3 = slightly, 2 = not at all, 1 = anti-religious)  
- **educ**: years of schooling  
- **ratemarr**: marital happiness (5 = very happy, 4 = happier than average, 3 = average, 2 = somewhat unhappy, 1 = very unhappy)  
- **naffairs**: number of affairs within last year  
- **affair**: =1 if had at least one affair  


Let's predict the number of affairs. Let's pick two variables and find the coefficients f:

\[
y_i=\beta_0+\beta_1x_{i3}+\beta_2x_{i2}+\epsilon_i
\]

Let's find the coefficients using the OLS formula:
$$\beta=(X'X)^{-1}X'y$$


```{r e1, exercise=TRUE}
#X=cbind(1, affairs$age, affairs$relig)
#y=affairs$naffairs  

```



```{r e2, exercise=TRUE}



```


Find predicted values

```{r e3, exercise=TRUE, exercise.setup="e2"}



```



Find residuals

```{r e4, exercise=TRUE, exercise.setup="e2"}



```

Make a prediction for particular values of x vector using beta vector from the first derivation.

```{r e5, exercise=TRUE, exercise.setup="e1"}



```


---

## Adding variables

Using dataset "card", regress wage (weekly, in cents) on educ (education in years). Then add IQ. How do the coefficient on education changes?



```{r e6, exercise=TRUE}



```




---

## Variance

We’ll work with data from a survey on extra-marital affairs in the US. Regress nafffairs on age and relig. Find the variance covaraince matrix. 


```{r e7, exercise=TRUE}



```

---

## Categorical variables in a regression

How does the marital status impacts earning?

Let's use cps data (survey  in the US). It provides information on both hourly wage (wagehr) and marital status (marstatus).  Possible options are: Married, Divorced, Widowed, Never married. 


```{r e8, exercise=TRUE}


```

---

## Interaction simple

Let's look at how value of the size of the house changes with its age. This data reports houses sold between 2006 and 2010. It has three variables of interest:

1. salesprice - sale price
2. lotarea - size of the lot
3. yearbuilt - year the house was built



```{r e9, exercise=TRUE}


```

---

## Interaction variable with a categorical variable with more categories

How do return to education vary across racial groups?

Let's use cps data (survey  in the US). It provides information on both hourly wage (wagehr) and racial group (race: Black, White, Other)  and education in years (educ).  

```{r 10, exercise=TRUE}

```


---

## R squared

Data "attend" reports final score of students of microeconomics from a US universities. It has following variables:

• attend: classes attended out of 32
• termGPA: GPA for term (GPA: grade point average)
• priGPA: cumulative GPA prior to term
• ACT: ACT score (standarized end of high school exam)
• final: final exam score
• atndrte: percent classes attended
• hwrte: percent homework turned in
• frosh: =1 if freshman
• soph: =1 if sophomore
• missed: number of classes missed


Outcome is final exam score. It goes from 0 to 40. (that's what we want to predict). Keep adding regressors and try to find the highest adjusted R square possible. (ex: attend, ACT, termGPA, priGPA)

```{r 11, exercise=TRUE}


```


---

## R squared

Data "attend" reports final score of students of microeconomics from a US universities. It has following variables:

• attend: classes attended out of 32
• termGPA: GPA for term (GPA: grade point average)
• priGPA: cumulative GPA prior to term
• ACT: ACT score (standarized end of high school exam)
• final: final exam score
• atndrte: percent classes attended
• hwrte: percent homework turned in
• frosh: =1 if freshman
• soph: =1 if sophomore
• missed: number of classes missed


Outcome is final exam score. It goes from 0 to 40. (that's what we want to predict). Keep adding regressors and try to find the highest adjusted R square possible. (ex: attend, ACT, termGPA, priGPA)

```{r 12, exercise=TRUE}


```


---

## Hypothesis testing

In the dataset "alcohol" we find information whether someone is employed (employ), whether they abuse alcohol (abuse), and years of education (educ).
Test if abusing alcohol affects your employment status when controlling for education. Compare to a model that doesn't include education as control. 

```{r 13f, exercise=TRUE}


##what is the distribution of test statistic? What is the number of degrees of freedom?

```


---

## Hypothesis testing with Interaction

Does education plays larger role if living in a big city?

The dataset "beauty" has information on wage (wage), years of education (educ), and whether someone lives in a big city (bigcity). Test if return to education is higher when you live in big city.

```{r 13, exercise=TRUE}

##what is the distribution of test statistic? What is the number of degrees of freedom?

```


---

## Confidence interval for a coefficient

The dataset sleep75 contains informatiomn on total hours of sleep (slpnaps), age, being married (marr), and hourly wage (hrwage).
Run a regression explaining sleep time using all these variables. What is 95% confidence interval on the impact of more dollar of hourly wage? What about being married?


```{r 14, exercise=TRUE}



##what is the distribution? What is the number of degrees of freedom?

```


---

## Confidence interval for a mean response


The dataset econmath contains data from a course in economics from Michigan State University. It reports the following variables. 

- score: final score
- age: age in years
- study: hours studying per week
- act: score on final high school exam
- fathcoll: 1 if father has college degree
- mothcoll: 1 if mother has college degree
- male: 1 if male

Let's find the confidence interval for the mean score for students who are 22 years old, study 10 hours per week, had 25 on final high school exam, only mother has a college degree, and is male

Step by step method:
1. Find the prediction
2. Write down the vector of predictors
3. Find the inverse of X'X
4. Find the standard error
5. Compute the standard errors of the prediction
6. Find critical value
7. Put together the interval

```{r 15m1, exercise=TRUE}


```

Quick method: 

```{r 15m2, exercise=TRUE}

lm_model <- lm(score ~ age+study+act+fathcoll+mothcoll+male, data = econmath)
summary(lm_model)
new_data<- data.frame(age=22, study=10, act=25, fathcoll=0, mothcoll=1, male=1)
predict(lm_model, newdata = new_data, interval = "confidence", level = 0.95, se.fit=TRUE)

```

---

## Confidence interval for a new observation

Let's use the same setting as before, but calculate the confidence interval for a new observation. 

Step by step method:
1. Find the prediction
2. Write down the vector of predictors
3. Find the inverse of X'X
4. Find the standard error
5. Compute the standard errors of the prediction
6. Find critical value
7. Put together the interval

```{r 16m1, exercise=TRUE}



```

Quick method: 

```{r 16m2, exercise=TRUE}

lm_model <- lm(score ~ age+study+act+fathcoll+mothcoll+male, data = econmath)
summary(lm_model)
new_data<- data.frame(age=22, study=10, act=25, fathcoll=0, mothcoll=1, male=1)
predict(lm_model, newdata = new_data, interval = "prediction", level = 0.95, se.fit=TRUE)

```


---

## ANOVA - Significance of the regression

The dataset discrim contains information of prices of fast food items from the US. Each observation is fast food restaurant.

Suppose you regress price of price of fries (pfries) on location characteristics: proportion black in zipcode (prpblck), proportion living in poverty (prppov), median family income (income). Does this regression helps at all to explain the price? 

1. Start with running the regression and looking at the results

```{r 17m1, exercise=TRUE}

lm_model <- lm(pfries ~ prpblck+prppov+income, data = discrim)
summary(lm_model)

```


2. Run the ANOVA test. 

```{r 17m2, exercise=TRUE, exercise.setup="17m1"}

anova(lm_model)
```


3. What is the SSE,  SST and SSR? What are their respective degrees of freedom? 

```{r 17m3, exercise=TRUE}


```


4. What is the F test result? What is the p-value?

```{r 17m4, exercise=TRUE}


#cumulative F distribution pf(q, df1, df2)



```


---

## Testing for difference in coefficients

Dataset married provides information on married couples in the US. It reports total family income and characteristics of both spouses: age (age_w, age_h), years of education (educ_w, educ_h), and body mass index (bmi_w, bmi_h).

Test if the impact of wife education is the same as the impact of husband education on family income.

1. Run the regression

```{r 18m1, exercise=TRUE}

a=lm(famincome~educ_w+educ_h, data=married)
summary(a)

```


2. Find the test statistic

```{r 18m2, exercise=TRUE, exercise.setup="18m1"}

vcov(a) #variance covariance matrix. 


```

3. Find the p-value

```{r 18m3, exercise=TRUE, exercise.setup="18m2"}


```


#### Alternative test. 

Now try again but with bmi of wife (bmi_w) vs bmi of husband (bmi_h)


```{r 18m5, exercise=TRUE}

a=lm(famincome~bmi_w+bmi_h, data=married)
summary(a)

```

---
