---
title: 'Class 4a: Simple Linear Regression'
author: "Business Forecasting"
output:
  xaringan::moon_reader:
    self_contained: true
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: true
      
---   
<style type="text/css">
.remark-slide-content {
    font-size: 20px;
}


</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dpi=300)
library(shiny)
library(ggplot2)
library(forecast)
library(plotly)
library(dplyr)
library(igraph)
library(reshape)
library(spData)
library(leaflet)
library(readr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(hrbrthemes)
library(gridExtra)
library(cowplot)
library(viridis)
library(gapminder)
library(knitr)
library(kableExtra)
library(DT)

```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", 
colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  blue = "#1E90FF",
  white = "#FFFFFF"
))
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
library(MASS) # for truehist function
```
---
## Roadmap


### This set of classes
- What is a multiple linear regression


---

Motivation example


AGE
 (clave de edad)
GENDER
Hora_ingreso?
MOTATE 0 motivo de urgencia
Number of other people at the urent care at the same facility

Outcome - tiempo estancia


show how the data looks like

---

Then show the equation and explains how it relates to the example

-interpretation - holding x2 constant!

---

Give another example on what could be the equation - or ask them.

Show an estimated example 



---

simulate data with 2 variables and one outcome

plot it on the 3d graphs (or do an app where they have to guess again)


---


Now imagine many more variable 
- can't vizualize it on the plot like that, but there will be other ways to do it. 



---

How do we estimate the parameters?

The same idea, we fit the line

We want to choose them so to minimize the estimation error


---
 
Minimization problem 

and solviing multiple equations

---

Easier to solve in a matrix form

Show in the matrix form

I give you x'x and x'y and calculate beta 

---

error estimation
hat matrix, error matrix


---

tos] Sea el modelo de regresi´on lineal m´ultiple dado por
y = Xβ + 
con los supuestos de de Gauss-Markov, entonces muestre que
yˆ
0
yˆ = βˆ0X0
y


---

Show with an example how it works, matrix by matrix 

---

or just plug into R


---


geometrical interpration


---

---

properties of oLS

unbiased, variance, covariance

---

INdicator dummy, categorical variables

- interpretation!

13. Para analizar la discriminaci´on salarial en una empresa se estim´o el siguiente modelo con 42 observaciones:
ln(SALi) = 1.5 − 0.08ln(EDi) − 0.05ln(EXi) + 0.03Hi + 0.02Pi + 0.03HiPi
donde SALi es el salario del trabajador, EDi y EXi su educaci´on y experiencia en a˜nos, respectivamente,
Hi una variable dicot´omica que toma el valor de 1 si el individuo es hombre y 0 si es mujer, y Pi una
variable dicot´omica que toma el valor de 1 si el individuo posee un posgrado y 0 si no.
a) Si adem´as se sabe que todos los coeficientes estimados son estad´ısticamente distintos de cero,
interprete el ajuste en t´erminos de si existe presencia de discriminaci´on salarial por g´enero al
tener distintos grados de estudio.
b) Determine el nivel de salario medio que esperar´a que tenga una mujer con 10 a˜nos de educaci´on,
5 de experiencia y que no posee un posgrado.



---

Interactions?

---

Model with squared variable (and powers)

---

It's all about partial correlation, ou cannot look at correlations separately!
  - correlation with the error, controlling for something and its effect

So for instance salary on some correlated variables. Once I keep constant one of them, the second one is not going to change much! 


---


r squared


---

Hypothesis testing/confidence intervals, same as before. easy once we know variance and expectation



---
test for single coefficient


---

test for significance of the regression
F-test or ANOVa... does it include beta0 or not?

testing (subsets) multiple linear restrictions

example


---

testing equality of coefficients
- can't be multicolinear

---

CI for regression coefficients


CI for mean response

CI for prediction of new observations

---
Comparing coefficients - standarized regression coefficients

