---
title: "LearnR Tutorial: Data Types and Visualizations"
author: "Business Forecasting"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(tidyverse)
library(DT)
library(MASS)
load("Health_data.Rda")
```

# 📊 Introduction

Welcome to this interactive **LearnR** tutorial on **data types** and **visualizations**. In this hands-on lesson, you will:

- Learn to distinguish between data types.
- Explore a real health survey dataset.
- Create and customize visualizations.
- Practice classifying variables interactively.


## 🧮 Exercise 0: Explore the Data Table

Let's start by viewing the dataset. This helps you understand the variables and their structure.

```{r ex0_table, exercise=TRUE}
# Show the first 10 rows of Health_data
head(Health_data, 10)
```


## 🧠 Exercise 1: Categorical vs Numerical Variables

Now try identifying whether a variable is categorical or numerical. We'll start with 'Gender'.

```{r ex1_question}
question("Is 'Gender' categorical or numerical?",
  answer("Categorical", correct = TRUE),
  answer("Numerical", correct = FALSE),
  allow_retry = TRUE
)
```


## 📈 Exercise 2: Frequency Table for Categorical Data

We often summarize categorical variables using frequency and proportion tables. Let’s see how often each value of `location_type` occurs.

```{r ex2_table, exercise=TRUE}
# Frequency and proportion of a categorical variable
table(Health_data$location_type)

prop.table(table(Health_data$location_type))
```


## 📊 Exercise 3: Bar Plot of Categorical Data

Bar plots are commonly used to visualize counts for categorical variables. Here we create one for `Difficulty_walking`.

```{r ex3_barplot, exercise=TRUE}
# Basic bar plot of Difficulty_walking
ggplot(data = Health_data, aes(x = Difficulty_walking)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(title = "Bar Plot of Difficulty Walking", x = "Difficulty Walking", y = "Count") +
  theme_minimal()
```


## 📊 Exercise 4: Bar Plot with Relative Frequencies

To better compare categories, we can display **proportions** instead of raw counts. This plot uses relative frequencies.

```{r ex4_barplot_prop, exercise=TRUE}
# Relative frequency bar plot
ggplot(Health_data, aes(x = Difficulty_walking)) +
  geom_bar(aes(y = ..prop.., group = 1), fill = "lightblue", color = "black") +
  labs(title = "Relative Frequency of Difficulty Walking",
       x = "Difficulty Walking", y = "Proportion") +
  theme_minimal()
```


## 🥧 Exercise 5: Pie Chart of Categorical Data

Another way to visualize proportions is with a pie chart. Let’s create one for `Difficulty_walking`.

```{r ex5_pie, exercise=TRUE}
# Pie chart of Difficulty_walking
ggplot(Health_data, aes(x = "", fill = Difficulty_walking)) +
  geom_bar(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Proportion of Difficulty Walking Responses", x = NULL, y = NULL) +
  theme_void() +
  theme(legend.title = element_blank())
```


## 📊 Exercise 6: What About Age?

Let’s examine the `age` variable. How many unique values are there? This helps determine whether it's categorical or continuous.

```{r ex6_age_levels, exercise=TRUE}

```


## 🧾 Exercise 7: Age as a Pie Chart 

Although we can create a pie chart for numeric variables, it usually isn't a good idea. Try it and see why.

```{r ex7_pie_age, exercise=TRUE}
# Pie chart of age — usually not recommended for numeric variables
ggplot(Health_data, aes(x = "", fill = factor(age))) +
  geom_bar(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Age Distribution (as Pie)", x = NULL, y = NULL) +
  theme_void() +
  theme(legend.title = element_blank())
```


## 📉 Exercise 8: Histogram of a Numerical Variable

Histograms are a better way to visualize the distribution of numerical variables. Let’s create one for `weight`.

```{r ex8_histogram, exercise=TRUE}
# Histogram of weight
ggplot(Health_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  labs(title = "Histogram of Weight", x = "Weight", y = "Frequency") +
  theme_minimal()
```


## 📦 Exercise 9: Summary Statistics and Boxplot

Let’s look at some key percentiles and visualize the distribution of `weight` with a boxplot.

```{r ex9_boxplot, exercise=TRUE}
# Summary statistics

quantile(Health_data$weight, c(0.7))

# Boxplot
ggplot(Health_data, aes(x = "", y = weight)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Box Plot of Weight", x = NULL, y = "Weight") +
  theme_minimal()
```

Let’s explore how quantiles divide a distribution. Move the slider to select a quantile. The histogram will update to highlight the proportion of data below the selected cutoff.

```{r ex9_ui, context="render"}
sliderInput("quantile_val", "Choose a Quantile:",
            min = 0.0001, max = 1, value = 0.5, step = 0.0001)
```

```{r ex9_quantile_plot, context="render"}
plotOutput("quantile_hist")
```

```{r ex9_quantile_server, context="server"}
output$quantile_hist <- renderPlot({
  q <- input$quantile_val
  x <- Health_data$weight
  q_val <- quantile(x, probs = q, na.rm = TRUE)
  df <- data.frame(x = x)

  ggplot(df, aes(x = x)) +
    geom_histogram(aes(y = ..density..,
                       fill = after_stat(x <= q_val)),
                   bins = 40, color = "white") +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey80"), guide = "none") +
    geom_vline(xintercept = q_val, color = "red", linetype = "dashed", linewidth = 1) +
    annotate("text", x = q_val, y = 0.02,
             label = paste0("q = ", round(q, 2), "\nvalue = ", round(q_val, 1)),
             vjust = -1, hjust = 0, color = "red", size = 4) +
    labs(title = "Weight Distribution with Quantile Cutoff",
         x = "Weight", y = "Density") +
    theme_minimal()
})
```


## 📦 Contingency table

Contingency tables help summarize the relationship between two categorical variables. Here, we examine whether having a mother with diabetes is associated with having diabetes oneself.
```{r ex10_data, exercise=TRUE}
# let's look at frequency of tables
table(Health_data$Mother_diabetes)
table(Health_data$diabetes) 
```

```{r ex10_contingency, exercise=TRUE}
# Create a contingency table between Mother's diabetes and respondent's diabetes status
table(Health_data$Mother_diabetes, Health_data$diabetes) #rows is whether mom has diabetes, and columns is whether respondent has diabetes
```

```{r ex10_prop, exercise=TRUE}
# View proportions within the entire sample
prop.table(table(Health_data$Mother_diabetes, Health_data$diabetes))
```

```{r ex10_prop_row, exercise=TRUE}
# View row-wise proportions (conditional on Mother_diabetes status)
prop.table(table(Health_data$Mother_diabetes, Health_data$diabetes), margin = 1)
```


## 🧪 Interactive Exploration: Covariance, Correlation, and Transformations

We’ll simulate variables \(X\) and \(Y\), then explore how adding or scaling affects their **covariance** and **correlation**.

---

### 🔧 Choose Parameters

```{r ui_covariance, context="render", echo=FALSE}
fluidPage(
  numericInput("mu_x", "Mean of X:", value = 80),
  numericInput("mu_y", "Mean of Y:", value = 50),
  numericInput("var_x", "Variance of X:", value = 225, min = 1),
  numericInput("var_y", "Variance of Y:", value = 100, min = 1),
  sliderInput("cov_xy", "Covariance between X and Y:", min = -150, max = 150, value = 100, step = 10),
  numericInput("add_const", "Add constant to X:", value = 0),
  numericInput("scale_x", "Scale factor for X:", value = 1),
  numericInput("scale_y", "Scale factor for Y:", value = 1)
)
```

---

### 📊 Output: Scatterplot and Summary

```{r plot_covariance, context="server", echo=FALSE}
output$plot_covariance <- renderPlot({
  library(MASS)
  set.seed(123)

  mu <- c(input$mu_x, input$mu_y)
  Sigma <- matrix(c(input$var_x, input$cov_xy,
                    input$cov_xy, input$var_y), nrow = 2, byrow = TRUE)

  data <- as.data.frame(mvrnorm(n = 1000, mu = mu, Sigma = Sigma))
  names(data) <- c("X", "Y")

  data$X_trans <- data$X * input$scale_x + input$add_const
  data$Y_trans <- data$Y * input$scale_y

  cov_val <- cov(data$X_trans, data$Y_trans)
  cor_val <- cor(data$X_trans, data$Y_trans)

  ggplot(data, aes(x = X_trans, y = Y_trans)) +
    geom_point(alpha = 0.3) +
    labs(
      title = paste0("Cov = ", round(cov_val, 2),
                     " | Cor = ", round(cor_val, 2)),
      x = "Transformed X", y = "Transformed Y"
    ) +
    theme_minimal()+
    xlim(0,150)+ylim(0,150)
})
```

```{r render_plot_covariance, context="render", echo=FALSE}
plotOutput("plot_covariance")
```

---

### 📉 Output: Histogram of X + Y

```{r hist_covariance, context="server", echo=FALSE}
output$hist_covariance <- renderPlot({
  library(MASS)
  set.seed(123)

  mu <- c(input$mu_x, input$mu_y)
  Sigma <- matrix(c(input$var_x, input$cov_xy,
                    input$cov_xy, input$var_y), nrow = 2, byrow = TRUE)

  data <- as.data.frame(mvrnorm(n = 1000, mu = mu, Sigma = Sigma))
  names(data) <- c("X", "Y")

  data$X_trans <- data$X * input$scale_x + input$add_const
  data$Y_trans <- data$Y * input$scale_y

  data$sum <- data$X_trans + data$Y_trans

  ggplot(data, aes(x = sum)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "white") +
    labs(title = "Histogram of Transformed X + Y", x = "X + Y", y = "Frequency") +
    theme_minimal()
})
```

```{r render_hist_covariance, context="render", echo=FALSE}
plotOutput("hist_covariance")
```