---
title: 'Class 2d: Review of concepts in Probability and Statistics'
author: "Business Forecasting"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncremdata:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAABIAAAASCAYAAABWzo5XAAAAWElEQVR42mNgGPTAxsZmJsVqQApgmGw1yApwKcQiT7phRBuCzzCSDSHGMKINIeDNmWQlA2IigKJwIssQkHdINgxfmBBtGDEBS3KCxBc7pMQgMYE5c/AXPwAwSX4lV3pTWwAAAABJRU5ErkJggg==entalSlides: true
      fig.retina: 2
---   
<style type="text/css">
.remark-slide-content {
    font-size: 20px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi=300)
library(shiny)
library(ggplot2)
library(forecast)
library(plotly)
library(dplyr)
library(igraph)
library(reshape)
library(spData)
library(leaflet)
library(readr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(hrbrthemes)
library(viridis)
library(reshape2)
library(gapminder)
library(knitr)
library(kableExtra)
library(DT)
library(patchwork)
options(scipen = 999)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", 
colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  blue = "#1E90FF",
  white = "#FFFFFF"
))
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
load("Data/Health_data.Rda")

```

---

layout: false
class: inverse, middle


# Summarizing Data 
## Comparisions and Associations



---

##Comparisions

- Descriptive and visual comparisons

--
- NOT declaring statistically significant differences, just eyeballing

--
- That's coming next


---

###Comparing categorical variables

####Does mother's diabetes predict diabetes?

- We have two categorical variables
- We can use frequency table to see how diabetes is distributed among groups with healthy vs diabetic mother

```{r 1, out.width='100%', fig.height=3, warning=FALSE}

Health_data$diabetes <- factor(Health_data$diabetes, levels = c(0, 1), labels = c("No", "Yes"))
Health_data$Mother_diabetes <- factor(Health_data$Mother_diabetes, levels = c(0, 1), labels = c("No", "Yes"))

# Create a frequency table
frequency_table <- table( Health_data$Mother_diabetes,Health_data$diabetes)

# Add variable names to row and column names
col_names <- c( "No Diabetes","Has Diabetes")
row_names <- c( "Mother No Diabetes","Mother Has Diabetes")
rownames(frequency_table) <- row_names
colnames(frequency_table) <- col_names


table_output <- kable(frequency_table, format = "html", caption = "") %>%
  kable_styling("striped", full_width = FALSE) 

table_output 
```

---
###Comparing categorical variables

####Does mother's diabetes predict diabetes?

- Are relative frequencies more helpful?
- Share of each subgroup within the sample

```{r 2, out.width='100%', fig.height=3, warning=FALSE, echo=FALSE, message=FALSE, results = 'hide'}

relative_frequency_table = round(frequency_table/nrow(Health_data),2)

# Add row and column margins to the relative frequency table
relative_frequency_table_with_margins <- addmargins(relative_frequency_table, FUN = list("Total" = sum))


# Use kable() to create a table with relative frequencies and margins
table_output <- kable(relative_frequency_table_with_margins, format = "html", caption = "", digits=2) %>%
  kable_styling("striped", full_width = FALSE)


```

```{r 2a, out.width='100%', fig.height=3, warning=FALSE, echo=FALSE, message=FALSE}

table_output
```

--

- Can we compare numbers in the *Has Diabetes* column?

--
- **Marginal frequencies** are total probabilities by group


---


#### Table of frequency


- We want to compare whether someone with diabetic mother is more likely to have diabetes than someone with healthy mother

--
- So we want to see whether: $$ \scriptsize P(Diabetes_i=1|\text{Mother has diabetes}_i=1)>P(Diabetes_i=1|\text{Mother has diabetes}_i=0)$$

--
- We want to look at the **relative conditional frequencies**
- They are usually in **contingency tables**
  - Share with diabetes within group of people whose mothers have diabetes
  - Share with diabetes within group of people whose mothers are healthy


--
```{r 2c, out.width='100%', fig.height=3, warning=FALSE, echo=FALSE, message=FALSE}

frequency_table <- table( Health_data$Mother_diabetes,Health_data$diabetes)

# Add variable names to row and column names
col_names <- c( "No Diabetes","Has Diabetes")
row_names <- c( "Mother No Diabetes","Mother Has Diabetes")
rownames(frequency_table) <- row_names
colnames(frequency_table) <- col_names

# Calculate relative frequencies
total_rows <- rowSums(frequency_table)
prop_table <- frequency_table / total_rows

table_output <-kable(prop_table, format = "html", caption = "", digits=2) %>%
  kable_styling("striped", full_width = FALSE)
table_output
```



--
- What about marginal frequencies here?
  - Row sums should add up to 1
      - $\scriptsize P(Diabetes_i=1|\text{Mother has diabetes}_i=1)+P(Diabetes_i=0|\text{Mother has diabetes}_i=1)$
  - Column sums are meaningless
      - $\scriptsize P(Diabetes_i=1|\text{Mother has diabetes}_i=1)+P(Diabetes_i=1|\text{Mother has diabetes}_i=0)$



---
- We can visualize it on a barplot

.center[
```{r 2d, out.width='100%', fig.height=4, warning=FALSE, echo=FALSE, message=FALSE}

summary_table <- table(Health_data$Mother_diabetes, Health_data$diabetes)
prop_table <- prop.table(summary_table, margin = 1)
prop_df <- as.data.frame(prop_table)

# Renaming columns for better labels
colnames(prop_df) <- c("Mother Has Diabetes", "Has Diabetes", "Proportion")


# Reshaping the dataframe for plotting



# Creating the bar plot using ggplot2
p=ggplot(prop_df, aes(x = `Has Diabetes`, y = Proportion, fill = `Mother Has Diabetes`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Has Diabetes", y = "Proportion", fill = "Mother has Diabetes") +
  theme(legend.position = "top")+
  theme_xaringan()+scale_xaringan_fill_discrete()
ggplotly(p,
        width = 800,   # Adjust the width according to your preference
        height = 400)
```
]
---

- Or better on a **stacked barplot**

.center[
```{r 2e, out.width='100%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

summary_table <- table(Health_data$Mother_diabetes, Health_data$diabetes)
prop_table <- prop.table(summary_table, margin = 1)
prop_df <- as.data.frame(prop_table)

# Renaming columns for better labels
colnames(prop_df) <- c("Mother Has Diabetes", "Has Diabetes", "Proportion")


# Reshaping the dataframe for plotting



# Creating the bar plot using ggplot2
p=ggplot(prop_df, aes(fill = `Has Diabetes`, y = Proportion, x = `Mother Has Diabetes`)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Mother has Diabetes", y = "Proportion", fill = "Has Diabetes") +
  theme(legend.position = "top")+
  theme_xaringan()+scale_xaringan_fill_discrete()

ggplotly(p,
        width = 800,   # Adjust the width according to your preference
        height = 400)
```
]

- *Stacked barplot* clearly shows the distribution of diabetes within each group

- Does it mean that having diabetic mother **causes** higher change of having diebetes?
---

### Comparing quantitative variables
#### Discrete variables


- For quantitative variables we can compare some summary statistics
 - *Example* means in two subpopulations
 - Are people older with diabetes older than people without it?


--
```{r 2f, out.width='100%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

mean_age_by_location <- Health_data %>%
  group_by(diabetes) %>%
  summarize(mean_age = mean(age))

# Reshaping the dataframe for plotting



# Creating the bar plot using ggplot2
p=ggplot(mean_age_by_location, aes(x = diabetes, y = mean_age, fill = diabetes)) +
  geom_bar(stat = "identity") +
  labs(title = "",
       x = "Diabetes",
       y = "Mean Age") +
  theme_xaringan()+scale_xaringan_fill_discrete()

ggplotly(p,
        width = 600,   # Adjust the width according to your preference
        height = 400)
```

---

### Comparing quantitative variables
#### Discrete variables

- Or we can do Box and Whiskers plots as before
- Or we can compare the whole distributions of frequencies

```{r 2g, out.width='100%', fig.height=3.8, warning=FALSE, echo=FALSE, message=FALSE}

ggplot(Health_data, aes(x = age, fill = factor(diabetes))) +
  geom_bar() +
  facet_wrap(~ diabetes, ncol = 1, 
             labeller = labeller(diabetes = c("No" = "No Diabetes", "Yes" = "Diabetes")),
             scales = "free_y") +
  labs(x = "Age",
       y = "Frequency") +
  theme_xaringan()+scale_xaringan_fill_discrete()+theme(legend.position="none")
```

---
#### Continous variables
- For continuous variables we can use the same methods (except frequency distribution) 
- Instead, we can compare densities or histograms
- Are men heavier than women?

--
.center[
```{r 2ha, out.width='100%', fig.height=3.8, warning=FALSE, echo=FALSE, message=FALSE}

ggplot(Health_data, aes(x = weight, fill=gender)) +
  geom_histogram(binwidth = 3, position = "identity", alpha = 0.7) +
  facet_wrap(~ gender, ncol = 1,
             scales = "free_y") +
  labs(
       x = "Age",
       y = "Count") +
  theme_xaringan()+scale_xaringan_fill_discrete()+theme(legend.position="none")


```
]


---
## Association


- Likely people would subscribe to the website to lose weight

--

- But do these people have resources?

--

- What is the relationship between Body Mass Index (BMI) and Income?

--

- More generally, how to measure .blue[association between two quantitative variables]

--

- Association between qualitative variables is measured with contingency tables 

```{r 2k, out.width='100%', fig.height=3.8, warning=FALSE, echo=FALSE, message=FALSE}

# Load required library
# Load required library
set.seed(123)

# Number of observations per city
n <- 200

# Create data for Mexico City
mexico_city <- data.frame(
  City = rep("Mexico City", n),
  BMI = round(rnorm(n, mean = 24, sd = 8),2),   # Mean BMI around 25, SD = 3
  Education = round(rnorm(n, mean = 12, sd = 2.5),1),  # Mean Education around 12 years, SD = 2
  Income = round(rnorm(n, mean = 180000, sd = 10000),2)  # Mean Income around 40000, SD = 10000
)

# Create data for Guadalajara
guadalajara <- data.frame(
  City = rep("Guadalajara", n),
  BMI = round(rnorm(n, mean = 26, sd = 8),2),   # Mean BMI around 25, SD = 3
  Education = round(rnorm(n, mean = 12, sd = 2.5),1),  # Mean Education around 12 years, SD = 2
  Income = round(rnorm(n, mean = 120000, sd = 8000),2)  # Mean Income around 60000, SD = 8000
)


# Add some positive correlation to the Income and Education variables in Guadalajara
guadalajara$Income <- guadalajara$Income - 5000 * guadalajara$BMI + 16000 * guadalajara$Education
mexico_city$Income <- mexico_city$Income - 2000 * mexico_city$BMI + 16000 * mexico_city$Education



# Combine the updated Guadalajara data with Mexico City data
dataset <- rbind(mexico_city, guadalajara)


```

---
### Associations

- Suppose we surveyed people from Guadalajara and CDMX about their .blue[BMI], .blue[education] and .blue[income]. 
- Scatter plots show associations between two quantitative variables
  - We put variables of interest (*example*: Y and X) on the axis
  - We place observation on the cartesian plane using their values of variable X and Y: $\{(x_1,y_1),(x_2,y_2)..\}$
- In our case: 
    - X axis is BMI
    - Y axis is Income
    - An individual $i$ is placed on these axis based on $(BMI_i, Income_i)$


```{r 2l, out.width='95%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}
datatable(dataset,
          fillContainer = FALSE,
          options = list(
            pageLength = 4,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE
)
```


---

```{r 2z, out.width='90%', fig.height=8, warning=FALSE, echo=FALSE, message=FALSE}
p=ggplot(data = dataset, aes(x = BMI, y = Income)) +
  geom_point() +
  labs(x = "BMI", y = "Income",
       color = "City") +
  theme_xaringan()+
   geom_segment(data = dataset[100, ],
               aes(xend = BMI, yend = Income, x = BMI, y = 0), 
               linetype = "dashed", color = "blue") +
  geom_segment(data = dataset[100, ],
               aes(xend = BMI, yend = Income, x = 0, y = Income), 
               linetype = "dashed", color = "blue") +
  geom_point(data = dataset[100, ],
             aes(x = BMI, y = Income), color = "red", size = 3) +
  geom_text(data = dataset[100, ],
            aes(label = paste("(", BMI, ",", Income, ")")), 
            vjust = -1, size=7) +
  theme(legend.position = "none")

ggplotly(p,
        width = 800,   # Adjust the width according to your preference
        height = 700)

```

---
### Assocations

- Would you say that the relationship is stronger in Guadalajara or in Mexico City?

```{r 2h, out.width='80%', fig.height=4, warning=FALSE, echo=FALSE, message=FALSE}
ggplot(data = dataset, aes(x = BMI, y = Income, color = City)) +
  geom_point() +
  labs(x = "BMI", y = "Income",
       color = "City") +
  theme_xaringan()+
  facet_wrap(~City)+
  theme(legend.position = "none")
```

- How to measure the strength of the relationship?

---
### Associations
#### Covariance

- **Covariance** measures the strength of the relationship between two variables. 

$$\text{Cov}(X, Y) = \frac{1}{N} \sum_{i=1}^{N} (x_i - \mu_X)(y_i - \mu_Y)$$

And it's sample equivalent is: 

$$\hat{\text{Cov}}(X, Y) = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})$$
--
- Covariance whether the two variables move together

--
- Covariance increases when:
  - The relationship is stronger
  - The deviations of variables are larger


---

`r knitr::include_url('https://shiny.rit.albany.edu/stat/rectangles/', height='550px')`

.footmark[
  Source: [https://shiny.rit.albany.edu/stat/rectangles/](https://shiny.rit.albany.edu/stat/rectangles/)
]

---
### Covariance

```{r 2zz, out.width='100%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

# Calculate covariances for each city
cov_mexico <- cov(dataset$BMI[dataset$City == "Mexico City"], dataset$Income[dataset$City == "Mexico City"])
cov_guadalajara <- cov(dataset$BMI[dataset$City == "Guadalajara"], dataset$Income[dataset$City == "Guadalajara"])

# Create individual plots with annotations
plot_mexico <- ggplot(data = subset(dataset, City == "Mexico City"), aes(x = BMI, y = Income)) +
  geom_point() +
  labs(x = "BMI CDMX", y = "Income") +
  annotate("text", x = 27, y = 0, label = paste("Cov:", round(cov_mexico, 2)), color = "blue")+
  theme_xaringan()+
  xlim(0,50)+ylim(-100000,400000)
  

plot_guadalajara <- ggplot(data = subset(dataset, City == "Guadalajara"), aes(x = BMI, y = Income)) +
  geom_point() +
  labs(x = "BMI GDL", y = "") +
  annotate("text", x = 27, y = 0, label = paste("Cov:", round(cov_guadalajara, 2)), color = "red")+
  theme_xaringan()+
  xlim(0,50)+ylim(-100000,400000)

# Combine the plots using patchwork
combined_plot <- plot_mexico + plot_guadalajara +
  plot_layout(ncol = 2, widths = c(1, 1.2))

# Display the combined plot
print(combined_plot)
```

---
### Covariance

- What has stronger relationship with Income: BMI or Years of Education?

```{r 2zzza, out.width='100%', fig.height=3, warning=FALSE, echo=FALSE, message=FALSE}

# Calculate covariances for each city
cov_mexico <- cov(dataset$Education, dataset$Income)
cov_guadalajara <- cov(dataset$BMI, dataset$Income)

# Create individual plots with annotations for covariances
plot_education <- ggplot(data = dataset, aes(x = Education, y = Income)) +
  geom_point() +
  labs(x = "Yrs of Educ.", y = "Income") +
  geom_text(data = data.frame(x = 30, y = 0, label = paste("Cov (Education):", round(cov_mexico, 2))), aes(x, y, label = label), color = "blue") +
  theme_xaringan()+
  xlim(0,50)

plot_bmi <- ggplot(data = dataset, aes(x = BMI, y = Income)) +
  geom_point() +
  labs(x = "BMI", y = "") +
  geom_text(data = data.frame(x = 30, y = 0, label = paste("Cov (BMI):", round(cov_guadalajara, 2))), aes(x, y, label = label), color = "red")+
  theme_xaringan()+
  xlim(0,50)

# Combine the plots using patchwork
combined_plot <- plot_education + plot_bmi +
  plot_layout(ncol = 2, widths = c(1, 1))

# Display the combined plot
print(combined_plot)

```

- BMI has larger covariance

--
- But we can't compare covariances of different variables

--
- Covariance depends on the scales (or units) of the variable

--
- All else equal, larger standard deviation implies larger covariance
  - The squares are just bigger

---

### Correlation

- **Correlation measures** the strength of a linear relationship between two variables.
- It ranges between -1 and 1 

**Population Correlation coefficient**:
$$\rho(X, Y) = \frac{\text{Cov}(X, Y)}{\sigma_X \cdot \sigma_Y}$$


**Sample Correlation coefficient**:

$$\hat{\rho}(X, Y) = \frac{\hat{\text{Cov}}(X, Y)}{s_X \cdot s_Y}$$
Where $s_X = \sqrt{\frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2}$


---
### Correlation

- Correlation is preferred over covariance because it's **scale-independent** and easier to interpret.
- Suppose that instead of measuring income (Y variable) in MXN , we measure it in Dollars. 
 - $Z$ income in dollars $Z=\frac{Y}{16}$
 
--
- Is $Cov(X,Z)=Cov(X,Y)$?

\begin{align*}
cov(X,Z) &=\frac{1}{N} \sum_{i=1}^{N} (x_i - \mu_X)(z_i - \mu_Z) \\
&=\frac{1}{N} \sum_{i=1}^{N} (x_i - \mu_X)(\frac{y_i}{16}- \frac{\mu_Y}{16}) \\
&=\frac{1}{16}\frac{1}{N} \sum_{i=1}^{N} (x_i - \mu_X)(y_i- \mu_Y) \\
& \neq cov(X,Y)
\end{align*}


---
### Correlation

- Correlation is preferred over covariance because it's **scale-independent** and easier to interpret.

- Suppose that instead of measuring income (Y variable) in MXN , we measure it in Dollars. 
 - $Z$ income in dollars $Z=\frac{Y}{16}$
 
--
- Is $\rho(X,Z)=\rho(X,Y)$?



\begin{align*}
\rho(X,Z) &=\frac{\frac{1}{N} \sum_{i=1}^{N} (x_i - \mu_X)(z_i - \mu_Z))}{\sqrt{\sum_{i=1}^{N} (x_i - \mu_X)^2} \cdot \sqrt{\sum_{i=1}^{N} (z_i - \mu_Z)^2}} \\
&=\frac{\frac{1}{N} \sum_{i=1}^{N} \sum_{i=1}^{N} (x_i - \mu_X)(\frac{y_i}{16}- \frac{\mu_Y}{16})}{\sqrt{\sum_{i=1}^{N} (x_i - \mu_X)^2} \cdot \sqrt{\sum_{i=1}^{N} (\frac{y_i}{16} - \frac{\mu_Y}{16})^2}} \\
&=\frac{\frac{1}{16} \frac{1}{N} \sum_{i=1}^{N} \sum_{i=1}^{N} (x_i - \mu_X)(y_i- \mu_Y)}{\frac{1}{16} \sqrt{\sum_{i=1}^{N} (x_i - \mu_X)^2} \cdot \sqrt{\sum_{i=1}^{N} (y_i - \mu_Y)^2}} \\
& = \rho(X,Y)
\end{align*}


---
### Correlation

- Correlation with education is actually stronger

```{r 2zzz, out.width='100%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

# Calculate covariances for each city
cov_mexico <- cor(dataset$Education, dataset$Income)
cov_guadalajara <- cor(dataset$BMI, dataset$Income)

# Create individual plots with annotations for covariances
plot_education <- ggplot(data = dataset, aes(x = Education, y = Income)) +
  geom_point() +
  labs(x = "Yrs of Educ.", y = "Income") +
  geom_text(data = data.frame(x = 30, y = -100000, label = paste("Cor (Education):", round(cov_mexico, 2))), aes(x, y, label = label), color = "blue") +
  theme_xaringan()+
  xlim(0,50)

plot_bmi <- ggplot(data = dataset, aes(x = BMI, y = Income)) +
  geom_point() +
  labs(x = "BMI", y = "") +
  geom_text(data = data.frame(x = 30, y = -100000, label = paste("Cor (BMI):", round(cov_guadalajara, 2))), aes(x, y, label = label), color = "red")+
  theme_xaringan()+
  xlim(0,50)

# Combine the plots using patchwork
combined_plot <- plot_education + plot_bmi +
  plot_layout(ncol = 2, widths = c(1, 1))

# Display the combined plot
print(combined_plot)

```



---
```{r 2zzzz, out.width='100%', fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}

library(plotly)

# Set random seed for reproducibility
set.seed(123)

# Generate data
n <- 100

xlim <- c(-4, 4)
ylim <- c(-8, 8)

# Positive correlation
x1 <- rnorm(n)
y1 <- x1 + rnorm(n)
plot1 <- plot_ly(x = x1, y = y1, text = paste("correlation:",round(cor(x1, y1), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

# Negative correlation
x2 <- rnorm(n)
y2 <- -x2 + rnorm(n)
plot2 <- plot_ly(x = x2, y = y2, text = paste("correlation:", round(cor(x2, y2), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

# No correlation
x3 <- rnorm(n)
y3 <- rnorm(n)
plot3 <- plot_ly(x = x3, y = y3, text = paste("correlation:",round(cor(x3, y3), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

x_base <- rnorm(n)
y_base <- x_base + rnorm(n)

cor_base <- cor(x_base, y_base)

# Create the first plot with slope 2 by scaling the base relationship
x4 <- x_base
y4 <- 2 * y_base
plot4 <- plot_ly(x = x4, y = y4, text = paste("correlation:",round(cor(x4, y4), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

# Create the second plot with slope 0.5 by scaling the base relationship
x5 <- 2 * x_base
y5 <- y_base
plot5 <- plot_ly(x = x5, y = y5, text = paste("correlation:",round(cor(x5, y5), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

# Non-linear relationship with near zero correlation
x6 <- seq(-2*pi, 2*pi, length.out = n)
y6 <- sin(x6) + rnorm(n, sd=0.5)
plot6 <- plot_ly(x = x6, y = y6, text = paste("correlation:",round(cor(x6, y6), 2)), type = 'scatter', mode = 'markers', showlegend = FALSE)%>%
layout(xaxis = list(range = xlim), yaxis = list(range = ylim))

# Negative correlation

# Combine plots into a single plot with multiple subplots (in 2 rows)
subplot(plot1, plot2, plot3, plot4, plot5, plot6, nrows = 2, margin = 0.02)

```

---
### Correlation

1. Correlation is a value between -1 and 1: $-1 \leq \rho(X, Y) \leq 1$.

--
2. Perfect positive correlation: $\rho = 1$. Perfect negative correlation: $\rho = -1$.

--
3. No linear correlation: $\rho = 0$, but this doesn't imply independence.

--
4. Correlation measures **linear** relationships; nonlinear relationships might not be accurately captured.

--
5. Correlation doesn't imply causation; a relationship could be coincidental.

