---
title: 'Class 2c: Review of concepts in Probability and Statistics'
author: "Business Forecasting"
output:
  xaringan::moon_reader:
    self_contained: true
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncremdataSlides: true
      fig.retina: 2
---   
<style type="text/css">
.remark-slide-content {
    font-size: 20px;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi=300)
library(shiny)
library(ggplot2)
library(forecast)
library(plotly)
library(dplyr)
library(gridExtra)
library(igraph)
library(reshape)
library(spData)
library(leaflet)
library(ggdist)
library(readr)
library(ggplot2)
library(tidyverse)
library(cowplot)
library(dplyr)
library(xaringan)
library(gridExtra)
library(dplyr)
library(hrbrthemes)
library(viridis)
library(gapminder)
library(knitr)
library(kableExtra)
library(DT)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", 
colors = c(
  red = "#f34213",
  purple = "#3e2f5b",
  orange = "#ff8811",
  green = "#136f63",
  blue = "#1E90FF",
  white = "#FFFFFF"
))
```


```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
load("Data/Health_data.Rda")

```

---

layout: false
class: inverse, middle


# Summarizing Data 
## Summary Statistics


---
# Measures of Dispersion


--

- Suppose a store has an average daily revenue of 10 000 pesos
- It could be that on each day it has exactly 10 000 pesos revenue
- Or it could be that on half of days it gets 20 000 pesos but on other half it gets 0 pesos
- Dispersion makes a big difference, especially when trying to understand risk! 


--
## Range

- **Range** the difference between minimum and maximum value in the data 


$$R=x_{max}-x_{min}$$


--

- What is the difference between the oldest and the youngest person with diabetes? 
- **R**=`r max(Health_data[Health_data$diabetes==1,]$age)-min(Health_data[Health_data$diabetes==1,]$age)`=`r format(round(max(Health_data[Health_data$diabetes==1,]$age),2))`-`r format(round(min(Health_data[Health_data$diabetes==1,]$age),2))`


---
- Very sensitive to outliers


```{r 11, out.width='100%', fig.height=5, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end

# Generate normal distributed data with mean 40 and sd 10
set.seed(123)  # for reproducibility
num_values <- 50
values <- rnorm(num_values, mean = 40, sd = 10)
data_initial <- data.frame(Values = values)

# Calculate range for initial dataset
range_initial <- range(values)

plot_initial <-  ggplot(data_initial, aes(x = Values)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = range_initial[1], y = 0.7, x = range_initial[2], yend = 0.7), color = "blue", size = 1.5) +
  geom_text(aes(x = round(c(range_initial[2] - range_initial[1]),2), y = 0.75, label = paste("Range: ", round(c(range_initial[2] - range_initial[1]),2))), 
            color = "blue", size = 6, vjust = -1) +
  labs(title = "", x = "Values", y = "Density")+theme_xaringan()+
  xlim(0,110)

# Modify one observation
modified_values <- c(values[-which.max(values)], 100)  # Replace last value with an outlier
data_modified <- data.frame(Values = modified_values)

# Calculate range for modified dataset
range_modified <- range(modified_values)

plot_modified <-  ggplot(data_modified, aes(x = Values)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = range_modified[1], y = 0.7, x = range_modified[2], yend = 0.7), color = "blue", size = 1.5) +
  geom_text(aes(x = round(range_modified[2] - range_modified[1],2)), y = 0.75, label = paste("Range: ", round(range_modified[2] - range_modified[1],2)), 
            color = "blue", size = 6, vjust = -1) +
  labs(title = "", x = "Values", y = "Density")+theme_xaringan()+
  xlim(0,110)


combined_plot <- plot_grid(plot_initial, plot_modified, labels = c("A", "B"), nrow = 2)

# Print the combined plot
print(combined_plot)


```

---
## Interquartile Range
- **Interquartile range** is the difference between the first and the third quartile of the data:
$$IQR=q_3-q_1$$
- What is the IQR of age in people with diabetes? 
- **IQR**=`r quantile(Health_data[Health_data$diabetes==1,]$age,0.75)-quantile(Health_data[Health_data$diabetes==1,]$age, 0.25)`=`r format(round(quantile(Health_data[Health_data$diabetes==1,]$age, 0.75),2))`-`r format(round(quantile(Health_data[Health_data$diabetes==1,]$age, 0.25),2))`

--
- 50% of the sample is between $q_3$ and $q_1$

--
```{r 2, out.width='100%', fig.height=3, warning=FALSE}
### Show two dotplots which have same mean/median but different interquartile ranges 

diabetes_data <- Health_data[Health_data$diabetes == 1, ]

# Calculate range and interquartile range (IQR)
range_diabetes <- range(diabetes_data$age)
iqr_diabetes <- IQR(diabetes_data$age)

plot_diabetes <- ggplot(diabetes_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black", show.legend = FALSE) +
  geom_segment(aes(xend = range_diabetes[1], y = 8, x = range_diabetes[2], yend = 8), color = "blue", size = 1.5) +
  geom_text(aes(x = mean(range_diabetes), y = 9, label = paste("Range: ", round(range_diabetes[2] - range_diabetes[1], 2))), 
            color = "blue", size = 6, vjust = -1) +
  geom_segment(aes(xend = quantile(diabetes_data$age, 0.25), y = 300, x = quantile(diabetes_data$age, 0.75), yend = 300), color = "red", size = 1.5) +
  geom_text(aes(x = quantile(diabetes_data$age, 0.5), y = 301, label = paste("IQR: ", round(iqr_diabetes, 2))), 
            color = "red", size = 6, vjust = -1) +
  labs(title = "", x = "Age", y = "Frequency") + 
  theme_xaringan() + xlim(0, max(diabetes_data$age) + 10)

# Print the plot
print(plot_diabetes)




```




---
- Is it more or less sensitive to outliers than range?

```{r 1, out.width='100%', fig.height=5, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end


# Generate normal distributed data with mean 40 and sd 10
set.seed(123)  # for reproducibility
num_values <- 50
values <- rnorm(num_values, mean = 40, sd = 10)
data_initial <- data.frame(Values = values)

# Calculate interquartile range (IQR) for initial dataset
iqr_initial <- IQR(values)

plot_initial <-  ggplot(data_initial, aes(x = Values)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = quantile(values, 0.25), y = 0.7, x = quantile(values, 0.75), yend = 0.7), color = "blue", size = 1.5) +
  geom_text(aes(x = quantile(values, 0.5), y = 0.75, label = paste("IQR: ", round(iqr_initial, 2))), 
            color = "blue", size = 6, vjust = -1) +
  labs(title = "", x = "Values", y = "Density") + theme_xaringan() +
  xlim(0, 110)

# Modify one observation
modified_values <- c(values[-which.max(values)], 100)  # Replace last value with an outlier
data_modified <- data.frame(Values = modified_values)

# Calculate interquartile range (IQR) for modified dataset
iqr_modified <- IQR(modified_values)

plot_modified <-  ggplot(data_modified, aes(x = Values)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = quantile(modified_values, 0.25), y = 0.7, x = quantile(modified_values, 0.75), yend = 0.7), color = "blue", size = 1.5) +
  geom_text(aes(x = quantile(modified_values, 0.5), y = 0.75, label = paste("IQR: ", round(iqr_modified, 2))), 
            color = "blue", size = 6, vjust = -1) +
  labs(title = "", x = "Values", y = "Density") + theme_xaringan() +
  xlim(0, 110)

combined_plot <- plot_grid(plot_initial, plot_modified, labels = c("A", "B"), nrow = 2)

# Print the combined plot
print(combined_plot)

```

---
### Interquartile Range 

---


### Example with data

- What is the IQR?

```{r, out.width='100%', fig.height=4, warning=FALSE}
# Set the seed for reproducibility
set.seed(42)

# Number of observations
num_observations <- 20

# Generate synthetic data for views (lognormal distribution with long right tail)
views <- round(rlnorm(num_observations, meanlog = 3, sdlog = 0.3))

# Generate video names
video_names <- paste("TikTok Video", 1:num_observations)

# Create the dataset
tiktok_dataset <- data.frame(VideoTitle = video_names, Views = views)
datatable(tiktok_dataset,
          fillContainer = FALSE,
          options = list(
            pageLength = 4,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE)

```



---

### Example with data

Here is a (smaller) data on distribution of how many views have various tik-tok videos. 
- Suppose that all views triples and 1000 additional people viewed them as well
$$y_i=3x_i+1000$$
- What is new IQR?
```{r, out.width='100%', fig.height=4, warning=FALSE}
# Set the seed for reproducibility
# Create the dataset
tiktok_dataset <- data.frame(VideoTitle = video_names, OldViews = views, NewViews=views*3+1000)
datatable(tiktok_dataset,
          fillContainer = FALSE,
          options = list(
            pageLength = 4,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE)

```

---
### IQR


- Order of observations was not affected, so same observations correspond to the first and the third quartile

$$q^{New}_{1}=3q^{Old}_1+1000$$
$$q^{New}_{3}=3q^{Old}_3+1000$$

--
- And more generally, for $$y_i=bx_i+a$$ and $b>0$

$$v^y_p=bv^{x}_p+a$$
- if $b<0$ then the order reverses. 

--
- So what does it mean for IQR?

$$IQR^{New}=q^{New}_{3}-q^{New}_{1}=3q^{Old}_3-3q^{Old}_1=3*IQR^{Old}$$


---


```{r 5aa, out.width='100%', fig.height=5, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end


# Generate normal distributed data with mean 40 and sd 10

data_initial <- data.frame(VideoTitle = video_names, OldViews = views, NewViews=views*3+1000)

# Calculate interquartile range (IQR) for initial dataset
iqr_initial <- IQR(data_initial$OldViews)

plot_initial <-  ggplot(data_initial, aes(x = OldViews)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = quantile(data_initial$OldViews, 0.25), y = 0.35, x = quantile(data_initial$OldViews, 0.75), yend = 0.35), color = "blue", size = 1.5) +
  labs(title = "", x = "Values", y = "Density") + theme_xaringan() +
  xlim(0,125)

# Calculate interquartile range (IQR) for modified dataset
iqr_modified <- IQR(data_initial$NewViews)

plot_modified <-  ggplot(data_initial, aes(x = NewViews)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  geom_segment(aes(xend = quantile(data_initial$NewViews, 0.25), y = 0.35, x = quantile(data_initial$NewViews, 0.75), yend = 0.35), color = "blue", size = 1.5) +
  labs(title = "", x = "Values", y = "Density") + theme_xaringan()+
  xlim(1000,1125)

combined_plot <- plot_grid(plot_initial, plot_modified, labels = c("A", "B"), nrow = 2)

# Print the combined plot
print(combined_plot)

```


---

### Variance & Standard Deviation

**Variance** measures how far an average observation is from the mean:

- **Population variance**: 
$$\sigma^2=E[(X-\mu)^2]=E[X^2]-\mu^2 =\frac{1}{N} \sum^N_{i=1}(x_i-\mu)^2=\frac{1}{N} (\sum^N_{i=1}x^2_i-N\mu^2)$$
For Discrete Variables it can be: $\sigma^2=\sum_k P(X=k)(k-\mu)^2$, where k is any possible value that X can take. 

--
- But variance does not have the right units since it squares everything...

--
- **Population standard deviation** deviation:

$$\sigma=\sqrt{\frac{1}{N} \sum^N_{i=1}(x_i-\mu)^2}$$
---


### Variance & Standard Deviation


--
- Why do we first take squares and then take square root? 

--
- Can't we just do $\frac{1}{N}\sum^N_{i=1}(x_i-\mu)$?

--
- NO! Because $\sum^N_{i=1}(x_i-\mu)=0$

--

- Why don't we just do Mean Absolute Deviation?
 - $MAD=\sum^N_{i=1}|(x_i-\mu)|$
 
 
--
- MAD is not differentiable at 0 :( 
- You can use it, but it's a different measure and will give different numbers. Why?


--
- Variance puts more weight on far away observations. 
- It's a weighted average distance, where weights are distances themselves. 

--
$$\sigma^2=\sum^N_{i=1}(x_i-\mu)^2=\frac{1}{N}\sum^N_{i=1} \underbrace{|(x_i-\mu)|}_{Weight}*\underbrace{|(x_i-\mu)|}_{Distance}$$
- It makes it very sensitive to outliers, which get a lot of weight!
- Standard deviation retains this property


---

```{r 3, out.width='100%', fig.height=5, warning=FALSE}
### Show three histograms with different level of variances: one symmetric, one symmetric with wider tail, and other one bimodal with much mass on both ends of the distribution. note how different is the variance. Annotate each with the standard deviation. 
library(ggplot2)

# Generate data for the histograms
set.seed(123)
data_symmetric <- rnorm(1000, mean = 0, sd = 1)
data_wide_tail <- rnorm(1000, mean = 0, sd = 2)
data_bimodal <- c(rnorm(500, mean = -2, sd = 0.5), rnorm(500, mean = 2, sd = 0.5))

# Create a data frame for plotting
df <- data.frame(
  Category = rep(c("Bimodal","Wide Tail", "Symmetric" ), each = 1000),
  Value = c(data_bimodal,data_wide_tail, data_symmetric )
)

means <- c(mean(data_symmetric), mean(data_wide_tail), mean(data_bimodal))
sds <- c(sd(data_symmetric), sd(data_wide_tail), sd(data_bimodal))


# Create histograms with annotations
p <- ggplot(df, aes(x = Value, fill = Category)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.6) +
  facet_grid(~ Category) +
  labs(x = "Value", y = "Frequency", fill = "Distribution Type") +
  theme_xaringan() +
  theme(legend.position = "none") +
 geom_vline(
    data = data.frame(Category = c("Symmetric", "Wide Tail", "Bimodal"),
                      Mean = means),
    aes(xintercept = Mean),
    linetype = "dashed",
    color = "red"
  ) 

print(p)


```

---

```{r 3b, out.width='100%', fig.height=5, warning=FALSE}
### Show three histograms with different level of variances: one symmetric, one symmetric with wider tail, and other one bimodal with much mass on both ends of the distribution. note how different is the variance. Annotate each with the standard deviation. 
library(ggplot2)

# Generate data for the histograms
set.seed(123)
data_symmetric <- rnorm(1000, mean = 0, sd = 1)
data_wide_tail <- rnorm(1000, mean = 0, sd = 2)
data_bimodal <- c(rnorm(500, mean = -2, sd = 0.5), rnorm(500, mean = 2, sd = 0.5))

# Create a data frame for plotting
df <- data.frame(
  Category = rep(c("Bimodal","Wide Tail", "Symmetric" ), each = 1000),
  Value = c(data_bimodal,data_wide_tail, data_symmetric )
)

means <- c(mean(data_symmetric), mean(data_wide_tail), mean(data_bimodal))
sds <- c(sd(data_symmetric), sd(data_wide_tail), sd(data_bimodal))


# Create histograms with annotations
p <- ggplot(df, aes(x = Value, fill = Category)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.6) +
  facet_grid(~ Category) +
  labs(x = "Value", y = "Frequency", fill = "Distribution Type") +
  theme_xaringan() +
  theme(legend.position = "none") +
  geom_text(
    data = data.frame(Category = c("Symmetric", "Wide Tail", "Bimodal"),
                      SD = c(sd(data_symmetric), sd(data_wide_tail), sd(data_bimodal))),
    aes(x = 0, y = 30, label = paste("SD =", round(SD, 2))),
    hjust = 0, vjust = 0,
    size = 5
  )+  geom_vline(
    data = data.frame(Category = c("Symmetric", "Wide Tail", "Bimodal"),
                      Mean = means),
    aes(xintercept = Mean),
    linetype = "dashed",
    color = "red"
  ) 

print(p)


```


---


### Variance & Standard Deviation

Consider two bets/situations:
- Bet A: with 75% you get 200 pesos and with 25% you lose me 200 pesos
  - 75% chance your life goes normal and you keep making money
  - 25% change your house burns down


--
- Bet B: with 75% you get 110 pesos and with 25% you get 70
  -  When your life goes normal you get 110 (you pay 90 for insurance)
  -  When your house burns down you are  paid some compensation (70)


--
- Compute expected value and variance of each bet
- Which one would you prefer?


--
- What if I change Bet B:
  - Bet B: with 75% you get 109 pesos and with 25% you get 69


--
- That's how insurance companies make profits


---

### Sample equivalents

- **Sample variance**: 
$$s^2=\frac{1}{n-1} \sum^n_{i=1}(x_i-\bar{x})^2=\frac{1}{n-1}( \sum^n_{i=1}x_i^2-n\bar{x}^2)$$
--
- **Sample standard deviation** deviation:

$$s=\sqrt{\frac{1}{n-1}( \sum^n_{i=1}x_i-\bar{x})^2}=\sqrt{\frac{1}{n-1}( \sum^n_{i=1}x_i^2-n\bar{x}^2)}$$
--
- Why we divide by $n-1$ rather than $n$?

--


---
- **Intuition** - observed values usually fall closer to the sample mean than to the population mean. Distances are artificially small. 

```{r 4, out.width='100%', fig.height=5, warning=FALSE}
### Show a vizualization illustrating this, population curve on the top (with mean marked), and below it three samples with their means marked 


# Generate data for the population
set.seed(123)
population_data <- rnorm(10000, mean = 0, sd = 1)

# Create data frame for plotting
population_df <- data.frame(Value = population_data)

# Calculate population mean
population_mean <- mean(population_data)

# Create the visualization
population_plot <- ggplot(population_df, aes(x = Value)) +
  geom_density(fill = "blue", alpha = 0.5) +
  geom_vline(xintercept = population_mean, linetype = "dashed", color = "black", linewidth=1.3) +
  labs(x = "Value", y = "Density") +
  theme_minimal()


# Generate samples from the population
num_samples <- 3
sample_size <- 20
samples <- lapply(1:num_samples, function(i) {
  sample_data <- sample(population_data, sample_size, replace = TRUE)
  data.frame(Sample = paste("Sample", i), Value = sample_data)
})


i=1
# Create a list of dotplot layers for the samples
dotplot_layers <- lapply(1:num_samples, function(i) {
  ggplot(samples[[i]], aes(x = Value)) +
    geom_dotplot(binwidth = 0.01, stackdir = "center",  dotsize = 10, alpha=0.7) +
    geom_vline(xintercept = mean(samples[[i]]$Value), linetype = "dashed", color = "red") +
    labs(x = NULL, y = NULL) +
    theme_minimal() +
    theme(legend.position = "none")+xlim(-4,4)
})

dotplot_layers_1=c(list(population_plot),dotplot_layers )


# Combine the population plot and the dotplot layers for the samples


arranged_plots <- grid.arrange(grobs = dotplot_layers_1, ncol = 1)


```

---
### Sample equivalents

- So the deviations from the sample mean underestimate the population standard deviation

--

- So we divide by a smaller number to correct for it

--

- In big sample $\frac{1}{n}$ and $\frac{1}{n-1}$ are similar, so correction doesn't matter as much



---
- **Intuition** - in big samples, our estimate of the population mean is already good, no need to correct

```{r 4b, out.width='100%', fig.height=5, warning=FALSE}
### Show a vizualization illustrating this, population curve on the top (with mean marked), and below it three samples with their means marked 


# Generate data for the population
set.seed(123)
population_data <- rnorm(10000, mean = 0, sd = 1)

# Create data frame for plotting
population_df <- data.frame(Value = population_data)

# Calculate population mean
population_mean <- mean(population_data)

# Create the visualization
population_plot <- ggplot(population_df, aes(x = Value)) +
  geom_density(fill = "blue", alpha = 0.5) +
  geom_vline(xintercept = population_mean, linetype = "dashed", color = "black", linewidth=1.3) +
  labs(x = "Value", y = "Density") +
  theme_minimal()


# Generate samples from the population
num_samples <- 3
sample_size <- 200
samples <- lapply(1:num_samples, function(i) {
  sample_data <- sample(population_data, sample_size, replace = TRUE)
  data.frame(Sample = paste("Sample", i), Value = sample_data)
})


i=1
# Create a list of dotplot layers for the samples
dotplot_layers <- lapply(1:num_samples, function(i) {
  ggplot(samples[[i]], aes(x = Value)) +
    geom_dotplot(binwidth = 0.01, stackdir = "center",  dotsize = 10, alpha=0.7) +
    geom_vline(xintercept = mean(samples[[i]]$Value), linetype = "dashed", color = "red") +
    labs(x = NULL, y = NULL) +
    theme_minimal() +
    theme(legend.position = "none")+xlim(-4,4)
})

dotplot_layers_1=c(list(population_plot),dotplot_layers )


# Combine the population plot and the dotplot layers for the samples


arranged_plots <- grid.arrange(grobs = dotplot_layers_1, ncol = 1)


```

---
### Standard Deviation

Consider a random variable X with $E(X)=\mu_x$ and standard deviation $\sigma_x$.

Ex: X is number of instagram followers distributed like this:

```{r 5aag, out.width='100%', fig.height=2, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end


# Generate normal distributed data with mean 40 and sd 10

data_initial <- data.frame(VideoTitle = video_names, OldViews = views, NewViews=views*3+1000)

# Calculate interquartile range (IQR) for initial dataset
iqr_initial <- IQR(data_initial$OldViews)

plot_initial <-  ggplot(data_initial, aes(x = OldViews)) +
  geom_dotplot(binwidth = 2, method = 'histodot', show.legend = FALSE) +
  labs(title = "", x = "Values", y = "Density") + theme_xaringan() +
  xlim(0,150)

# Print the combined plot
print(plot_initial)

```


---
### Standard Deviation

- What happens to mean and standard deviation if everyone gets 40 more followers?
- $Y=X+40$

```{r 5aagfb, out.width='100%', fig.height=2, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end


# Generate normal distributed data with mean 40 and sd 10

data_initial <- data.frame(VideoTitle = video_names, OldViews = views, NewViews=views+40)


plot_modified <- ggplot(data_initial) +
  # Original OldViews in gray
  geom_dotplot(aes(x = OldViews), binwidth = 2, method = 'histodot', fill = "gray", alpha = 0.4, color=NA, show.legend = FALSE) +
  # NewViews
  geom_dotplot(aes(x = NewViews), binwidth = 2, method = 'histodot', fill = "blue", show.legend = FALSE) +
  labs(title = "", x = "Values", y = "Density") +
  theme_xaringan() +
  xlim(0, 150)

plot_modified
```

--

$$\scriptsize \mu_Y=E(Y)=\frac{\sum_{i=1}^N(y_i+40)}{N}=\frac{\sum_{i=1}^N(x_i+40)}{N}=\frac{\sum_{i=1}^Nx_i+N*40}{N}=E(X)+40=\mu_X+40$$

--

$$\scriptsize \sigma_Y=\sqrt{Var(Y)}=\sqrt{\frac{\sum_i(y_i-\mu_y)^2}{N}}=\sqrt{\frac{\sum_i((x_i+40)-(\mu_x+40))^2}{N}}=\sqrt{\frac{\sum_i(x_i-\mu_x)^2}{N}}=\sqrt{Var(X)}=\sigma_X$$
$$\scriptsize E(X+c)=E(X)+c \quad\text{and} \quad Var(X+c)=Var(X)$$

---
### Standard Deviation

- What happens to mean and standard deviation if everyone followers get multiplied by 3? (without addition)
- $Y=3*X$

```{r 5aagf, out.width='100%', fig.height=2, warning=FALSE}
### Illustrate with two dotplots, where I just change one observation at the end


# Generate normal distributed data with mean 40 and sd 10

data_initial <- data.frame(VideoTitle = video_names, OldViews = views, NewViews=views*3)

plot_modified <- ggplot(data_initial) +
  # Original OldViews in gray
  geom_dotplot(aes(x = OldViews), binwidth = 2, method = 'histodot', fill = "gray", alpha = 0.4, color=NA, show.legend = FALSE) +
  # NewViews
  geom_dotplot(aes(x = NewViews), binwidth = 2, method = 'histodot', fill = "blue", show.legend = FALSE) +
  labs(title = "", x = "Values", y = "Density") +
  theme_xaringan() +
  xlim(0, 150)

plot_modified


```
--

$$\scriptsize \mu_Y=E(Y)=\frac{\sum_iy_i}{N}=\frac{\sum_i3x_i}{N}=3\frac{\sum_ix_i}{N}=3E(X)=3\mu_X$$

--
$$\scriptsize \sigma_Y=\sqrt{Var(Y)}=\sqrt{\frac{\sum_i(y_i-\mu_y)^2}{N}}=\sqrt{\frac{\sum_i(3x_i-3\mu_x)^2}{N}}=\sqrt{3^2\frac{\sum_i(x_i-\mu_x)^2}{N}}=\sqrt{3^2Var(X)}=3\sigma_X$$
--
$$\scriptsize E(cX)=cE(X) \quad\text{and} \quad Var(cX)=c^2Var(X)$$

---

### Coefficient of Variation

**Coefficient of Variation** divides the standard deviation by the mean. 

$$C.V.=\frac{\sigma}{|\mu|}$$

And sample eqiuvalent
$$c.v.=\frac{s}{|\bar{x}|}$$
--

- Why?
  - It expresses standard deviation as proportion of the mean
      - Small value means variation is low compared to the mean
  - It is unit free
  - You can compare it across variables with different units/magnitudes

---

### Coefficient of Variation

**Example** - variation of stocks in different currencies 

.pull-left[
```{r 6, out.width='100%', fig.height=3, warning=FALSE}
### Table with stocks in Pesos vs some Stock in Dollars 

library(dplyr)

# Set seed for reproducibility
set.seed(123)

# Generate date sequence
start_date <- as.Date("2023-07-01")
dates <- seq(start_date, by = "day", length.out = 20)

# Generate data for MXN and USD stocks
data <- data.frame(
  Date = dates,
  MXN_Stock = round(rnorm(20, mean = 100, sd = 15),2),
  USD_Stock = round(rnorm(20, mean = 1.2, sd = 0.18),2)
)

datatable(data,
          fillContainer = FALSE,
          options = list(
            pageLength = 6,
            searching = FALSE,
            initComplete = JS(
              "function(settings, json) {",
              "$(this.api().table().container()).css({'font-size': '12px'});",
              "}"
            )
          ),
          rownames = FALSE
)


```
]

.pull-right[
- **Standard deviation**:
  - USD: 0.149
  - MXN: 14.59


- **Coefficient of variation**:
  - USD: 0.12
  - MXN: 0.14
]
---

### Coefficient of Variation

So more generally, if $y_i=bx_i$, then

$$C.V._y=\frac{\sigma_y}{|\mu_y|}=\frac{|b|\sigma_x}{|b\mu_x|}=C.V._x$$
What if $y_i=bx_i+a$?

--

Then 

$$C.V._y=\frac{\sigma_y}{|\mu_y|}=\frac{|b|\sigma_x}{|b\mu_x+a|} \neq C.V._x$$
---
## Percentiles


<center>
<img src=Exam_q_CV.png width="800">
</center>



---

### Box and Whiskers plot 

- Helps to see the distribution of the data
- Helps to see to see the outliers
  - Outliers are useful to see anomalies and potential errors in data collection
  - Whisker can be maximally 1.5 times the interquartile range
  - Any point beyond that is an outlier
  - If no point beyond 1.5 times the interquartile range, whisker goes just to the last datapoint and is shorter than 1.5 times the interquartile range

```{r 7a, out.width='100%', fig.height=3, warning=FALSE}
### Box and whiskers plot for weight (using ggdist)

set.seed(123)
data <- data.frame(values = rnorm(1000))

combined_plot <- ggplot(data, aes(y=values)) +
  
  # Rotated and mirrored histogram on the left
  geom_histogram(aes(x = -..density.. * 0.5), bins=30, fill="skyblue", orientation="y") +
  
  # Boxplot on the right, slightly further from the histogram
  geom_boxplot(aes(x=0.1), fill="pink", width=0.05) +
  
  # Annotations for the boxplot closer to the box
  annotate("text", x = -0.04, y = quantile(data$values, 0.25), label = "Q1", hjust=0) +
  annotate("text", x = -0.04, y = median(data$values), label = "Median", hjust=0) +
  annotate("text", x = -0.04, y = quantile(data$values, 0.75), label = "Q3", hjust=0) +
  annotate("text", x = -0.04, y = max(data$values[data$values < (quantile(data$values, 0.75) + 1.5 * IQR(data$values))]), 
           label = expression(paste("Upper Whisker: Q3 + 1.5 * IQR")), hjust=0) +
  annotate("text", x = -0.04, y = min(data$values[data$values > (quantile(data$values, 0.25) - 1.5 * IQR(data$values))]), 
           label = expression(paste("Lower Whisker: Q1 - 1.5 * IQR")), hjust=0) +
  
  # Aesthetics and labels
  theme_minimal() +
  theme(axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(title = "")

# Displaying the plot
print(combined_plot)


```




---
### Box and Whiskers plot 

**Dataset comparisons**
- They summarize data very well
```{r 8a, out.width='100%', fig.height=4.5, warning=FALSE}
### Box and whiskiers plot for two different distributions, with different amount of variation and different amount of outliers
### wages in food sector vs wages in accounting sector 

Health_data %>% 
  ggplot(aes(x = factor(diabetes), y = age, fill = factor(diabetes))) +
  # add half-violin from {ggdist} package
  geom_boxplot(
    width = 0.12,
    # purple outlier points
    outlier.color = "purple",
    alpha = 0.5)+
  # boxplot
  coord_flip()+
  theme_xaringan()+
  theme(legend.position = "none")

```

---

### Box and Whiskers plot 

**Dataset comparisons**
- They summarize data very well
```{r 8b, out.width='100%', fig.height=4.5, warning=FALSE}
### Box and whiskiers plot for two different distributions, with different amount of variation and different amount of outliers
### wages in food sector vs wages in accounting sector 

Health_data %>% 
  ggplot(aes(x = factor(diabetes), y = age, fill = factor(diabetes))) +
  # add half-violin from {ggdist} package
  geom_boxplot(
    width = 0.12,
    # purple outlier points
    outlier.color = "purple",
    alpha = 0.5)+
  stat_halfeye(
    # adjust bandwidth
    adjust = 0.5,
    # move to the right
    justification = -0.2,
    # remove the slab interval
    .width = 0,
    point_colour = NA) +
  # boxplot
  coord_flip()+
  theme_xaringan() +
  theme(legend.position = "none")

```

---

<center>
<img src=Exam_q_bw.png width="800">
</center>


---
### Exercises:

- Review Exercises:
  - PDF 2: 1,2,6,8 (skip f),9,10,13,

- Homeworks
  - Lista 00.1: 1,2,4,5

