---
title: "LearnR Tutorial: Sampling and Estimation with Airbnb Prices"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(ggplot2)
library(dplyr)
load("listing_full_price.Rda")  # assumes it loads `listings` with column `price`
```

---

## 📦 Section 1: Explore One Sample

### ✏️ Choose a sample size and draw a sample

Use the widget below to set your sample size. Then write code to:

- Filter out invalid prices (e.g., NA, 0, too high)
- Draw a sample of `n` listings from the price data
- Show the distribution of the full data and of your sample
- Calculate the sample mean


```{r one_sample_exercise, exercise=TRUE}
# Filter the data
n=100 #choose sample size

# Draw one sample

my_sample <- sample(listings$price, size = n, replace = FALSE)

# Plot full data vs sample
par(mfrow = c(1, 2))
hist(listings$price, breaks = 30, main = "Full Population", col = "lightgray", xlab = "Price")
hist(my_sample, breaks = 15, main = paste("Sample (n =", n, ")"), col = "skyblue", xlab = "Sample Price")

# Show the sample mean
paste("Sample mean is",mean(my_sample,na.rm=TRUE))
```

---

## 📊 Section 2: Sampling Distribution (Shiny)

Explore the **sampling distribution** of the sample mean across many samples.

```{r sample_ui2, context="render"}
fluidPage(
  sliderInput("n2", "Sample Size", min = 5, max = 200, value = 30, step = 5),
  numericInput("reps", "Number of Replications", value = 1000, min = 100),
  actionButton("resample2", "Generate Sampling Distribution")
)
```

```{r show_sd, context="server"}
output$sd_prices <- renderText({
  paste0("Population SD of prices: ", round(sd(listings$price, na.rm = TRUE), 2))
})
```

```{r show_sd_output, context="render", echo=FALSE}
verbatimTextOutput("sd_prices")
```

Calculate the theoretical variance of the estimator

```{r two_sample_exercise, exercise=TRUE}

```

```{r sample_dist_logic, context="server"}
observeEvent(input$resample2, {
  prices <- listings$price

  set.seed(123)
  sample_means <- replicate(input$reps, {
    mean(sample(prices, size = input$n2, replace = FALSE))
  })

  true_mean <- mean(prices)

  output$sampdist_plot2 <- renderPlot({
    ggplot(data.frame(sample_means), aes(x = sample_means)) +
      geom_histogram(bins = 200, fill = "steelblue", color = "white") +
      geom_vline(xintercept = true_mean, col = "red", linetype = "dashed", linewidth = 1.2) +
      labs(title = "Sampling Distribution of the Sample Mean",
           subtitle = paste0("True Mean: ", round(true_mean, 2), " MXN"),
           x = "Sample Mean", y = "Frequency") +
      theme_minimal()+xlim(0,10000)
  })

  output$sampdist_stats2 <- renderText({
    paste0("Mean of sample means: ", round(mean(sample_means, na.rm=TRUE), 2), "\n",
           "True mean: ", round(mean(prices,na.rm=TRUE), 2), "\n",
           "SD of sample means: ", round(sd(sample_means, na.rm=TRUE), 2), "\n",
           "Theoretical SE: ", round(sd(prices, na.rm=TRUE) / sqrt(input$n2), 2))
  })
})
```

```{r sample_dist_output2, context="render"}
plotOutput("sampdist_plot2")
verbatimTextOutput("sampdist_stats2")
```



## 🧪 Bernoulli to Normal: Step-by-Step Visualization

### From 0/1 Decisions to Continuous Approximations

We simulate calls in Cancun and Puerto Vallarta:

- Each user makes a **Bernoulli decision**: call (1) or not (0)
- As we sample more users, the **sum** of Bernoullis becomes approximately normal
- The **sum of two normals** is also normal

---

```{r bern_ui, context="render", echo=FALSE}
fluidPage(
  sliderInput("n_cancun", "Number of users in Cancun", min = 1, max = 1000, value = 100, step = 10),
  sliderInput("p_cancun", "Prob(Call) in Cancun", min = 0, max = 1, value = 0.4, step = 0.01),
  sliderInput("n_pv", "Number of users in Puerto Vallarta", min = 1, max = 1000, value = 80, step = 10),
  sliderInput("p_pv", "Prob(Call) in Puerto Vallarta", min = 0, max = 1, value = 0.6, step = 0.01),
  numericInput("cutoff", "Cutoff: Total Calls >", value = 100, min = 0),
  checkboxInput("use_normal", "Use normal approximation", value = TRUE),
  actionButton("simulate", "Simulate and Show")
)
```

```{r bern_logic, context="server"}
observeEvent(input$simulate, {
  n1 <- input$n_cancun
  p1 <- input$p_cancun
  n2 <- input$n_pv
  p2 <- input$p_pv
  k <- input$cutoff

  reps <- 10000

  # Step 1: Simulate Bernoulli outcomes
  bern1 <- rbinom(reps, size = n1, prob = p1)
  bern2 <- rbinom(reps, size = n2, prob = p2)
  total <- bern1 + bern2

  mu_total <- n1 * p1 + n2 * p2
  sd_total <- sqrt(n1 * p1 * (1 - p1) + n2 * p2 * (1 - p2))

  output$bernoulli_plot <- renderPlot({
    par(mfrow = c(2, 2))

    barplot(c(1 - p1, p1), names.arg = c("0", "1"), main = "Bernoulli: Cancun", col = "lightblue")
    barplot(c(1 - p2, p2), names.arg = c("0", "1"), main = "Bernoulli: PV", col = "orange")

    hist(bern1, breaks = 40, main = paste0("Distribution of Sum of ", n1, " Bernoullis (Cancun)"),
         col = "lightblue", xlab = "# Calls", xlim= c(0, n1))
    hist(bern2, breaks = 40, main = paste0("Distribution of Sum of ", n2, " Bernoullis (PV)"),
         col = "orange", xlab = "# Calls", xlim= c(0, n2))

    par(mfrow = c(1, 1))
  })

output$total_plot <- renderPlot({
  df <- data.frame(total = total)
  g <- ggplot(df, aes(x = total)) +
    geom_histogram(aes(y = ..density..), bins = 100, fill = "purple", alpha = 0.7, color = "white") +
    labs(title = "Distribution of Total Calls (Cancun + PV)",
         x = "Total Calls", y = "Density") +
    theme_minimal()

  if (input$use_normal) {
    xseq <- seq(0, n1 + n2, length.out = 400)
    norm_density <- dnorm(xseq, mean = mu_total, sd = sd_total)
    g <- g + geom_line(data = data.frame(x = xseq, y = norm_density),
                       aes(x = x, y = y), color = "red", linewidth = 1.2)
  }

  g + geom_vline(xintercept = k, linetype = "dashed", color = "red") +
    annotate("text", x = k + 3, y = max(density(total)$y), label = paste0("Cutoff: ", k),
             hjust = 0, vjust = 1.5, color = "red") +
    xlim(0, n1 + n2)
})

  output$tail_prob <- renderText({
    if (input$use_normal) {
      prob <- pnorm(k, mean = mu_total, sd = sd_total, lower.tail = FALSE)
      paste0("Using normal approximation:\nP(total > ", k, ") ≈ ", round(prob, 4))
    } else {
      empirical <- mean(total > k)
      paste0("Using simulation:\nP(total > ", k, ") ≈ ", round(empirical, 4))
    }
  })
})
```

```{r bern_outputs, context="render", echo=FALSE}
plotOutput("bernoulli_plot")
plotOutput("total_plot")
verbatimTextOutput("tail_prob")
```

## 📊 Section 3: Comparing Sampling Distributions of Sample Variance

Explore how the sampling distribution of **sample variance** depends on the shape of the population. Compare the results from a **normal distribution** and a **skewed distribution** (exponential). We also overlay the theoretical chi-squared distribution scaled by the population variance.

```{r var_ui_outputs_compare, context="render", echo=FALSE}
fluidPage(
  sliderInput("n_var", "Sample Size", min = 5, max = 200, value = 20, step = 5),
  numericInput("reps_var", "Number of Replications", value = 10000, min = 100),
  actionButton("resample_var", "Generate Sampling Distributions"),
  h4("Sample Variance from Normal Distribution"),
  plotOutput("sampdist_plot_norm"),
  verbatimTextOutput("sampdist_stats_norm"),
  h4("Sample Variance from Non-Normal (Exponential) Distribution"),
  plotOutput("sampdist_plot_skewed"),
  verbatimTextOutput("sampdist_stats_skewed")
)
```

```{r var_server_logic_compare, context="server"}
observeEvent(input$resample_var, {
  set.seed(123)

  # Normal distribution
  population_norm <- rnorm(100000, mean = 50, sd = 20)
  true_var_norm <- var(population_norm)
  sample_vars_norm <- replicate(input$reps_var, {
    var(sample(population_norm, size = input$n_var, replace = FALSE))
  })

  output$sampdist_plot_norm <- renderPlot({
    df <- input$n_var - 1
    x_vals <- seq(0, quantile(sample_vars_norm, 0.999, na.rm = TRUE), length.out = 400)
    chi_scaled <- dchisq(x_vals / true_var_norm * df, df = df) * df / true_var_norm

    ggplot(data.frame(sample_vars = sample_vars_norm), aes(x = sample_vars)) +
      geom_histogram(aes(y = ..density..), bins = 60, fill = "steelblue", color = "white") +
      geom_line(data = data.frame(x = x_vals, y = chi_scaled), aes(x = x, y = y), color = "red", linewidth = 1) +
      geom_vline(xintercept = true_var_norm, col = "black", linetype = "dashed", linewidth = 1) +
      labs(title = "Normal Distribution", x = "Sample Variance", y = "Density") +
      theme_minimal()
  })


  # Exponential distribution
  population_skewed <- rexp(100000, rate = 1/50)
  true_var_skewed <- var(population_skewed)
  sample_vars_skewed <- replicate(input$reps_var, {
    var(sample(population_skewed, size = input$n_var, replace = FALSE))
  })

  output$sampdist_plot_skewed <- renderPlot({
    df <- input$n_var - 1
    x_vals <- seq(0, quantile(sample_vars_skewed, 0.999, na.rm = TRUE), length.out = 400)
    chi_scaled <- dchisq(x_vals / true_var_skewed * df, df = df) * df / true_var_skewed

    ggplot(data.frame(sample_vars = sample_vars_skewed), aes(x = sample_vars)) +
      geom_histogram(aes(y = ..density..), bins = 60, fill = "darkorange", color = "white") +
      geom_line(data = data.frame(x = x_vals, y = chi_scaled), aes(x = x, y = y), color = "red", linewidth = 1) +
      geom_vline(xintercept = true_var_skewed, col = "black", linetype = "dashed", linewidth = 1) +
      labs(title = "Exponential Distribution", x = "Sample Variance", y = "Density") +
      theme_minimal()
  })

})
```

## 📊 Visualizing the Connection (Interactive)

Adjust the cutoff to see how the shaded area under the standard normal corresponds to the right-tail of the chi-squared distribution.  
Also verify that:  
\[
P(Z^2 > c) = P(|Z| > \sqrt{c}) = P(Z < -\sqrt{c}) + P(Z > \sqrt{c})
\]

```{r ui_chi_visual, context="render", echo=FALSE}
sliderInput("cutoff", "Cutoff value (Z² > ?)", min = 0.1, max = 5, value = 1, step = 0.1)
```

```{r server_chi_visual, context="server"}
observe({
  cutoff <- input$cutoff
  set.seed(42)
  z <- rnorm(100000)
  x <- z^2

  # Breakpoint
  sqrt_c <- sqrt(cutoff)

  # Plot 1: Standard Normal with shaded tails where Z² > cutoff
  hist_z <- hist(z, breaks = 200, plot = FALSE)
  hist_z$counts <- hist_z$counts / sum(hist_z$counts)
  cols_z <- ifelse(hist_z$breaks < -sqrt_c | hist_z$breaks > sqrt_c, "purple", rgb(0.2, 0.2, 0.2, 0.2))

  # Plot 2: Chi-squared with X > cutoff
  hist_x <- hist(x, breaks = 200, plot = FALSE)
  hist_x$counts <- hist_x$counts / sum(hist_x$counts)
  cols_x <- ifelse(hist_x$breaks > cutoff, "purple", rgb(0.2, 0.2, 0.2, 0.2))

  # Numerical check: empirical and theoretical
  prob_empirical <- mean(x > cutoff)
  prob_equivalent <- mean(z < -sqrt_c | z > sqrt_c)
  prob_theoretical <- 2 * pnorm(-sqrt_c)

  output$chi_connection_plot <- renderPlot({
    par(mfrow = c(1, 2))
    plot(hist_z, col = cols_z, border = FALSE, main = "Standard Normal", ylab = "Density", xlab = "z", xlim = c(-4, 4))
    plot(hist_x, col = cols_x, border = FALSE, main = expression(Chi^2 * "(1)"), ylab = "Density", xlab = "x", xlim = c(0, 6))
    par(mfrow = c(1, 1))
  })

  output$chi_connection_test <- renderText({
    paste0(
      "Cutoff: Z² > ", round(cutoff, 2), "   (|Z| > ", round(sqrt_c, 2), ")\n\n",
      "Empirical P(Z² > c): ", round(prob_empirical, 4), "\n",
      "Empirical P(Z < -sqrt(c) or Z > sqrt(c)): ", round(prob_equivalent, 4), 
    )
  })
})
```

```{r render_chi_visual, context="render"}
plotOutput("chi_connection_plot")
verbatimTextOutput("chi_connection_test")
```